\section*{Class I Verbs}
\begin{itemize}
\item \anpl{baddbedd} \apl{baddnen} \defi{to dry up}
\begin{itemize}
\item \exsen{Ine da däbeddnegän, ada källäm a, ada walle mäg a, orbam a wa karama da ade petapeta abal gogon.} \extran{Water dried up, and the ponds, the big creeks, pools, and swamps became very shallow.}
\end{itemize}
\item \anpl{balldae} \apl{[]} \defi{to announce}
\item \anpl{bebe} \apl{benan} \defi{to leak, excrete}
\begin{itemize}
\item \exsen{Källa bebe we nalle.} \extran{[You] go and pass excrement.}
\item \exsen{De däräng a källa benanbenan dan.} \extran{This dog is always pooping.}
\item \exsen{Sana sisor a nyukukum me bebe allan.} \extran{The new sago is leaking in the sago bag.}
\end{itemize}
\item \anpl{bendoe} \apl{bendoenen} \defi{to confuse, mix up}
\begin{itemize}
\item \exsen{Ngäna ubim komlla angde ikop deyagän, ngäna yambedowaeyan.} \extran{When I saw them two, I confused them for each other.}
\end{itemize}
\item \anpl{bengae} \apl{bengaenen} \defi{to roof, cover, roof, roofing}
\begin{itemize}
\item \exsen{Ubi ttägäll de nyeny alle daibengaemallo.} \extran{They covered the mumus with nyeny wood.}
\end{itemize}
\item \anpl{bobllem} \apl{[]} \defi{to flap, shake (of skin)}
\begin{itemize}
\item \exsen{Mälla bo pätt a bobllem allan.} \extran{The skin of the woman's body shakes.}
\end{itemize}
\item \anpl{bollboll} \apl{bollnen} \defi{to open sago}
\begin{itemize}
\item \exsen{Bogo ako ttongo pättkäp däbollän gädagäde we.} \extran{He opened another node of the sago trunk to beat.}
\end{itemize}
\item \anpl{bäblem} \apl{[]} \defi{shiver}
\item \anpl{bäbälläd} \apl{bällädnen} \defi{to drop without warning}
\item \anpl{bällabälle} \apl{bällanan} \defi{to find}
\begin{itemize}
\item \exsen{Ngäna kumuddäga ttägäll käp tärpitärpi de nabllenegan.} \extran{I found three smooth stones.}
\end{itemize}
\item \anpl{bälämbäl} \apl{bälnan} \defi{to miss, long for, to remember, think of}
\begin{itemize}
\item \exsen{Ngäna ttongo lla de bälämbäl eran.} \extran{I miss someone.}
\item \exsen{Angde lla da borale bo eka de bäddareyo, ubi mɨnyi ngattongatt masarmasar de bämbälaebeyo.} \extran{When people hear the song of the flute, they will remember their ancestors from before.}
\item \exsen{Ngäma ddob gagäll gudae att gagäll de dämbälaebnalla.} \extran{Some of us (excl.) were remembering bad things from the past.}
\end{itemize}
\item \anpl{bämblläd} \apl{blländnan} \defi{to expand, rise}
\item \anpl{bänybäny} \apl{bänynan} \defi{to cut, slice (flesh)}
\begin{itemize}
\item \exsen{Ngämi ge ddia de däbänya.} \extran{We (excl.) cut this deer.}
\end{itemize}
\item \anpl{dadäräb} \apl{darbnan} \defi{to write, to dress, to decorate}
\begin{itemize}
\item \exsen{Ngämo kame dan dadräb a.} \extran{I can't write.}
\item \exsen{Pizin eka de ai dan badräb.} \extran{I can write in Tok Pisin.}
\item \exsen{Bogo obom dadäräb eran.} \extran{She is dressing her.}
\item \exsen{Bogo mɨnyi llo ttam alle godräballe ingong de ngkamalle bandra peyang.} \extran{She would decorate herself with leaves and start to dance to the song.}
\end{itemize}
\item \anpl{dadäräd} \apl{[]} \defi{to clear (e.g. of the mind, sky)}
\begin{itemize}
\item \exsen{Näkäp a odardan.} \extran{The mind is cleared.}
\end{itemize}
\item \anpl{ddaddell} \apl{[]} \defi{destroy}
\item \anpl{ddaddu} \apl{ddanun} \defi{to remove a shoot}
\item \anpl{ddaddällɨg} \apl{ddallɨgnen} \defi{to destroy}
\begin{itemize}
\item \exsen{Kuddäll lla da bäne ma de däddellɨgallo.} \extran{They destroy the house of the dead man.}
\end{itemize}
\item \anpl{ddangoe} \apl{ddangoenen} \defi{to force to do}
\begin{itemize}
\item \exsen{Andrew Yuga bom era tongoe e daddengoenän.} \extran{Andrew was forcing Yuga to play a game.}
\end{itemize}
\item \anpl{ddäddäg} \apl{ddägnan} \defi{to peel, remove}
\begin{itemize}
\item \exsen{Ngäna ma papek de näddägaban.} \extran{I removed the walls.}
\end{itemize}
\item \anpl{ddäddäg} \apl{ddägnan} \defi{to pain, ache, hurt}
\begin{itemize}
\item \exsen{Ge ute da naddägnan ngänäm.} \extran{This sore is hurting me.}
\end{itemize}
\item \anpl{ddäddäg} \apl{ddägnan} \defi{to look, watch, to look, watch}
\item \anpl{ddäddäg} \apl{ddägnan} \defi{to bind}
\begin{itemize}
\item \exsen{Sana dor däpittaemeya a däddägaemeya.} \extran{We wove and bound the sago stalks.}
\end{itemize}
\item \anpl{ddäddäl} \apl{[]} \defi{to shove}
\item \anpl{ddägaddäge} \apl{[]} \defi{to write, to branch out}
\begin{itemize}
\item \exsen{Amo ingoll anykeanyke da bin peyang goddgenegän gänya mani me?} \extran{Whose likeness and name are inscribed on this coin?}
\end{itemize}
\item \anpl{ddällgoe} \apl{[]} \defi{to go through thick bush in search of animals, to disturb a beehive}
\begin{itemize}
\item \exsen{Lla da däräng peyang tawa de daddällgoeneyo.} \extran{The man perused the swamp together with the dog.}
\end{itemize}
\item \anpl{ddänddäl} \apl{ddälnan} \defi{to climb}
\begin{itemize}
\item \exsen{Baet a llo ddage me gonddälän.} \extran{The cuscus climbed on the tree branch.}
\end{itemize}
\item \anpl{ddänddäm} \apl{ddämnen, ddämnan} \defi{to drown, to set (of the moon), to drown, sink, to worry, sink}
\begin{itemize}
\item \exsen{Obom dandämoyän ttu we, angde obom ttu we dandämoyän bogo ddone dängllawän, be bogo gonddämän.} \extran{She pushed him into the deep, and when she pushed him into the deep, he didn't swim; he drowned.}
\item \exsen{Kauga angde Matthew bom ikop dägagän ddänddäm me, bogo mängalae gogbänän walle we, Matthew bom dirängänän.} \extran{When Kauga saw Matthew drowning, he quickly jumped into the water and lifted Matthew out.}
\item \exsen{Kok a de ddänddäm allan.} \extran{The moon is setting.}
\item \exsen{Kollba da ddone dan, ede enda nett de nänddäman?} \extran{There are no fish, so what sank the net?}
\item \exsen{Bogo ereya gagäll de bällnan eran. Diba ttoen da obo näkäp de ddänddäm eran.} \extran{She's remembering bad news. This thing makes her worry.}
\end{itemize}
\item \anpl{ddänggaddängge} \apl{ddänganen} \defi{to crucify, to catch, trap}
\begin{itemize}
\item \exsen{Yesu bom llo tärpamatt toko me danddgeyo.} \extran{They crucified Jesus on a wooden cross.}
\item \exsen{Martin käza de danddgewän.} \extran{Martin trapped the crocodile.}
\end{itemize}
\item \anpl{dedre} \apl{dernan} \defi{to descend, go down}
\begin{itemize}
\item \exsen{Ubi tutu atta walle we godrenegnän.} \extran{They were descending from the mountain to the river.}
\end{itemize}
\item \anpl{dodo} \apl{donen} \defi{to blow over, only with plural objects, to crumble sago into frying pan, to remove thorn}
\item \anpl{dodro} \apl{dronen} \defi{to slip, to die}
\begin{itemize}
\item \exsen{Nikki bo kaptte da dronendronen dag.} \extran{Nikki's clothes are slipping off.}
\item \exsen{Kollba da tämamae dadrowän a be kottllam aebe ttam dagirnän.} \extran{The fish all died but the turtle lived on.}
\end{itemize}
\item \anpl{dodro} \apl{dronen} \defi{to clean}
\begin{itemize}
\item \exsen{Bogo llan de komlla deyadurowän mermerae.} \extran{She washed her two ears well.}
\end{itemize}
\item \anpl{domoe} \apl{[]} \defi{to push}
\begin{itemize}
\item \exsen{Ngäna gall de domoe eran.} \extran{I am pushing the boat.}
\end{itemize}
\item \anpl{dradre} \apl{[]} \defi{to dress}
\begin{itemize}
\item \exsen{Ngäna obom dädre.} \extran{I dressed him.}
\end{itemize}
\item \anpl{däbae} \apl{däbaenen} \defi{to drizzle, to knock fruit continuously}
\begin{itemize}
\item \exsen{Sisri abal yogoll a adäbaeyan.} \extran{It just started to drizzle.}
\end{itemize}
\item \anpl{dädik} \apl{däknen} \defi{to calm someone down with cool words from anger, to bump}
\item \anpl{dädäräb} \apl{däräbnan} \defi{to cut grass, flowers}
\item \anpl{dämoe} \apl{dämoenen} \defi{to push, to send}
\begin{itemize}
\item \exsen{Ngäna abo däbe sɨmell de misdae ada dandämoe.} \extran{Then I just pushed the pig like that.}
\item \exsen{Angde obom ttu wi dandämoeyän, bogo ddone dängllawän, be bogo gonddämän.} \extran{When she pushed him into the deep, he didn't swim: rather, he drowned.}
\item \exsen{Mälla de dandmoeaebeya wayati kongkom e.} \extran{We sent the women to fetch the watermelons.}
\item \exsen{Giniya män kälsre de nandämoeyan Bodog pate.} \extran{Giniya sent the girl to Bodog.}
\end{itemize}
\item \anpl{dändär} \apl{därmäll} \defi{to stuff}
\begin{itemize}
\item \exsen{Mänmän a sana de zazaba we dändäraebeyo.} \extran{The girls stuffed the sago into the bag.}
\end{itemize}
\item \anpl{dändärek} \apl{[]} \defi{to control, influence, rule, govern}
\begin{itemize}
\item \exsen{Bogo mɨnyi llɨg de gagäll ttoen ngasnges e bändärekän.} \extran{He will pressure the boy to do bad things.}
\end{itemize}
\item \anpl{däradära} \apl{däranen} \defi{to decorate}
\item \anpl{därmae} \apl{[]} \defi{to wipe out}
\begin{itemize}
\item \exsen{Bogenvil me lla da mäse därmae e gongkaemnegän.} \extran{In Bougainville, people (i.e. the PNG Defense Force) were trying to wipe them (i.e. the secessionists) out.}
\end{itemize}
\item \anpl{eddrugoe} \apl{[]} \defi{to shout}
\item \anpl{eka laem} \apl{laemnen} \defi{to argue}
\begin{itemize}
\item \exsen{Eka golaemeyo.} \extran{They argued.}
\end{itemize}
\item \anpl{eka panypeny} \apl{[]} \defi{to speak}
\begin{itemize}
\item \exsen{Ddone ngänäm mulldae allan eka panypeny e.} \extran{I am unable to speak.}
\item \exsen{Ge Llimollang a alla ingollang eka däpanyneyo.} \extran{This is how Limol people spoke it.}
\end{itemize}
\item \anpl{ellollo} \apl{ellonin} \defi{to put down load and rest}
\item \anpl{ergod} \apl{ergodnen} \defi{to crawl}
\begin{itemize}
\item \exsen{Llɨg kälsre da ergodnen allan.} \extran{The child is crawling.}
\end{itemize}
\item \anpl{erär} \apl{ernan} \defi{to name; pass down a name, to measure}
\begin{itemize}
\item \exsen{Ngäma pemli makäp me gänyan ge ernan erallo.} \extran{Here in our (excl.) family, we are passing on this name.}
\item \exsen{Ngämo kame dan aeya ngämo bin di ge daerän.} \extran{I don't know who gave me this name.}
\item \exsen{Ge za da llo ernan ma dan.} \extran{This thing is for measuring trees.}
\end{itemize}
\item \anpl{gag} \apl{[]} \defi{to store}
\begin{itemize}
\item \exsen{Mätta de gaguma we dägagaeballo.} \extran{They stored the yams in the yamhouse.}
\end{itemize}
\item \anpl{gagälläd} \apl{galldänen} \defi{to cut leaf}
\item \anpl{gallgell} \apl{gallnen} \defi{to remove leaves, take leaves off; remove skin, skin, to open mouth wide when screaming}
\item \anpl{gllae} \apl{gllaenen, gllagllae} \defi{to float, drift; paddle, to paddle; pedal, to dig, spade}
\begin{itemize}
\item \exsen{Net a gllaenen allan.} \extran{The net is floating.}
\item \exsen{Ngämi (excl.) angde gogllaenalla, däbe za da dibeya ka ili därän.} \extran{When we paddled, that thing almost disappeared.}
\item \exsen{Ngäna mɨnyi gall de bogllae.} \extran{I will paddle the canoe.}
\item \exsen{Ngämi kili gogman, kili peyang gall de dagllaeya do Upiara dälltam.} \extran{We (excl.) were so happy, with joy we paddled the canoe until we arrived at Upiara.}
\item \exsen{Ngäna gall de mängamängall dagllae ddia koenmäll e, gall alle gumbiebmeny.} \extran{I paddled the canoe quickly to chase the deer with the boat.}
\item \exsen{Gällae de dängkamalle, dagllaealle dagllaealle.} \extran{They started digging; they dug and dug.}
\end{itemize}
\item \anpl{gllae} \apl{[]} \defi{to shine}
\begin{itemize}
\item \exsen{Yäbäd a mermerae gogllaewän.} \extran{The sun shined nicely.}
\end{itemize}
\item \anpl{gllaglle} \apl{gllanen} \defi{to skin, remove skin}
\begin{itemize}
\item \exsen{Ngäna ttall ttoe de nägllewan.} \extran{I removed the wallaby skin.}
\end{itemize}
\item \anpl{gogo} \apl{gonen} \defi{to build}
\begin{itemize}
\item \exsen{Ma de gogo eran.} \extran{He is building the house.}
\item \exsen{Nogowallo.} \extran{They built it.}
\item \exsen{Ttongo täräp me ngäna ngämo ma gogo we gotäbane.} \extran{One time, I planned to build my house.}
\item \exsen{Abo bongo ttongo ma sisor de nogo.} \extran{You must build a new house.}
\end{itemize}
\item \anpl{gomoe} \apl{gomoenen} \defi{to make a mistake}
\begin{itemize}
\item \exsen{Ngäna gem eka de ttättle anggan, ge ddob lla da erem gomoenen amallo.} \extran{I correct these words, which other people mispronounce.}
\end{itemize}
\item \anpl{gonagone} \apl{gonanen} \defi{to cook, to burn, to heat}
\begin{itemize}
\item \exsen{Bongo nalle, otät e nognen.} \extran{You, go and cook the food.}
\item \exsen{Llo mit de dugonaeballo.} \extran{They burned the tree bases.}
\item \exsen{Ngäna ine ttämttämang yu dägag a ngämo ttäle de dogonaeya.} \extran{I boiled water and heated my leg.}
\end{itemize}
\item \anpl{gumae} \apl{[]} \defi{to go around}
\item \anpl{gädagäde} \apl{gädanen} \defi{to beat sago, pound sago, to sharpen, to throw a tantrum}
\begin{itemize}
\item \exsen{Beyawaebe dägdene.} \extran{Alone, he beat sago.}
\item \exsen{Ngäna tri pos dae dägdeneg.} \extran{I only sharpened three posts.}
\end{itemize}
\item \anpl{gäddgädd} \apl{gäddnan} \defi{to fight, to catch, to cut (leaves)}
\begin{itemize}
\item \exsen{Lla da gäddgädd e bobllabän.} \extran{The men go to fight.}
\item \exsen{Bongo sämell aba peyang gäddgädd alle.} \extran{You are fighting with the pig.}
\item \exsen{Ubi gogäddän.} \extran{They fought.}
\item \exsen{Mälla da sisri ag me kok de nägäddaeballo kollba tokong e.} \extran{This morning, the women caught grasshopers for bait.}
\item \exsen{Ngäma Dowabunang me gällall de ke ami dägäddaebeyo?} \extran{Who cut our (excl.) pandanus leaves in Dowabunang?}
\end{itemize}
\item \anpl{gäglib} \apl{glibnen} \defi{to chase, scare away/off (wild animals), to pluck}
\begin{itemize}
\item \exsen{Däräng a ddia de diglibiyu Madura pate.} \extran{The dogs chased the deer, steering it towards Madura.}
\item \exsen{Ngäna käza de nägliban.} \extran{I scared away the crocodile.}
\item \exsen{Pa kom de dägläbaebeyo, tot do dangesaemeyo.} \extran{They plucked the bird's feathers and made a mess there.}
\end{itemize}
\item \anpl{gäglläb} \apl{glläbnen} \defi{to weed for the first time (hard work to remove the large plants)}
\item \anpl{gämoe} \apl{gämoenen} \defi{to miss, to miss, feel longing for}
\begin{itemize}
\item \exsen{Ddia de ddone dagmoe bazu.} \extran{Shooting the deer, I didn't miss.}
\item \exsen{Ttongo eka nagmoeyan.} \extran{I forgot to say one thing.}
\item \exsen{Ngäna bam gämoe nallan.} \extran{I am missing you.}
\end{itemize}
\item \anpl{gängglläd} \apl{glländnan} \defi{to shove, push, to extend (e.g. a garden)}
\begin{itemize}
\item \exsen{Ngäna sɨmell de duduli dänggllädalle.} \extran{I shoved the pig that way.}
\end{itemize}
\item \anpl{gänggälläm} \apl{gällämnan} \defi{to wash (a body part or inanimate object)}
\begin{itemize}
\item \exsen{Ddob mälla da pane moepang de ikrol alle gällämnan amallo.} \extran{Some women wash dirty pots with ash.}
\item \exsen{Bogo pilatt de dängllämnegan.} \extran{He washed the dishes.}
\item \exsen{Da bongo källama atta angos, ttang anggälläm.} \extran{If you come back from the toilet, wash your hands.}
\end{itemize}
\item \anpl{gɨg} \apl{gɨnen} \defi{to collect ants}
\item \anpl{gɨlgɨl} \apl{[]} \defi{descend}
\item \anpl{i} \apl{inen} \defi{to weave; interlock}
\begin{itemize}
\item \exsen{Bogo nyäng de i eran.} \extran{She is weaving the basket.}
\item \exsen{Tater de aya diwän?} \extran{Who wove the mat?}
\item \exsen{Ngämo yae era lla ngänäm polle dinignän.} \extran{My mother, she used to build sturdy fences like a man.}
\end{itemize}
\item \anpl{ibae} \apl{[]} \defi{to dig using one's nose}
\item \anpl{ikop ddäddäg} \apl{[]} \defi{to look, watch, to look, watch}
\begin{itemize}
\item \exsen{Llɨg a wa män a ubi ikop goddägneyo.} \extran{The boy and the girl looked at each other.}
\item \exsen{Masar era ngämim ikop deyaddägnän.} \extran{Grandfather, he was watching us.}
\end{itemize}
\item \anpl{imonz} \apl{imonznen} \defi{to touch}
\item \anpl{imullgoe} \apl{[]} \defi{to drop, push, make fall}
\begin{itemize}
\item \exsen{Bogo obom daimullgoeyän.} \extran{She pushed him over.}
\end{itemize}
\item \anpl{inggoe} \apl{[]} \defi{shake}
\item \anpl{inungoe} \apl{inungoenen} \defi{to shake (when dancing)}
\begin{itemize}
\item \exsen{Bongo nainungoenalleǃ} \extran{You were shaking!}
\end{itemize}
\item \anpl{itrae} \apl{itraenen} \defi{to hang}
\item \anpl{ittɨtt} \apl{ittnen} \defi{to catch (an aquatic animal)}
\begin{itemize}
\item \exsen{Angde toto gognän, Kila a Wala gall alle tudi ittaenen e deyareyo bem e.} \extran{When evening came, Kila and Wala went to the sea to fish with the canoe.}
\item \exsen{Bogo komlla guzi de deyaittän.} \extran{She caught two crawfish.}
\item \exsen{Ttongdae kollba da guittän.} \extran{One fish was caught.}
\end{itemize}
\item \anpl{kaekep} \apl{kaepnen} \defi{to chew, to struggle to cut (e.g. with a dull knife)}
\begin{itemize}
\item \exsen{Gae da nge ingollae dan, be obo täma da kaekep ma dan.} \extran{The gae palm is like coconut, but its husk is for chewing.}
\end{itemize}
\item \anpl{kakällät} \apl{kalltänen} \defi{to split a leaf}
\item \anpl{kallkell} \apl{kallnen} \defi{to deleaf a plant to reveal the shoot, or to take the skin off}
\item \anpl{klloklloe} \apl{klloenen, klloklloe} \defi{to gather, join, come together, to gather, bring together, mix}
\begin{itemize}
\item \exsen{Ubi goklloeyän Malläm ttängäm e.} \extran{They gathered in Malam village.}
\item \exsen{Chairman a obo melem a ada dan ge, lla de baklloenegnän eka tameny e.} \extran{This is the chairman's job: he will gather the people for discussions.}
\item \exsen{Ngämi eka de klloklloe panynen eralla.} \extran{We are speaking mixing languages.}
\end{itemize}
\item \anpl{kokllo} \apl{kllonen} \defi{to scratch, scrape}
\begin{itemize}
\item \exsen{Ngäna nge de kokllo eran.} \extran{I'm scraping a coconut.}
\item \exsen{Lla ulle da bandrabandrog bun akllonan.} \extran{The big man was scratching his head while dancing.}
\end{itemize}
\item \anpl{koko} \apl{konen} \defi{to cut (flesh or meat), butcher}
\begin{itemize}
\item \exsen{Bogo ddäddäg de dokonggän oblle.} \extran{He cut the meat for her.}
\item \exsen{Ddia koko we ngäna llɨg di dokom.} \extran{I took the boys to butcher the deer.}
\end{itemize}
\item \anpl{kokop} \apl{kopnen} \defi{to peel, skin, husk}
\begin{itemize}
\item \exsen{Ge lla da nae de kokop eran.} \extran{This man is peeling a sweet potato.}
\item \exsen{Ttongo lla ulle da up o de näkopnan.} \extran{A big man peeled a ripe banana.}
\end{itemize}
\item \anpl{kokto} \apl{koktonen} \defi{to bail (water)}
\begin{itemize}
\item \exsen{Ngäna gall e godagal, ine dokto.} \extran{I got in the canoe and bailed the water out.}
\end{itemize}
\item \anpl{konakone} \apl{konanen} \defi{to cover, cover, sheet, blanket}
\begin{itemize}
\item \exsen{Yäbäd de ddapall käkan da dakonewän.} \extran{Clouds covered the sun.}
\item \exsen{Däbe män a konakone alle gokonewän.} \extran{That girl covered up with a blanket.}
\end{itemize}
\item \anpl{kunen} \apl{[]} \defi{to flee, scatter, run away}
\begin{itemize}
\item \exsen{Nikol angde walle menae e nallan, ddone ada kollba nukuaemallo.} \extran{When Nikol went to the side of the river, many fish scattered away.}
\end{itemize}
\item \anpl{kädbae} \apl{[]} \defi{to test, try}
\begin{itemize}
\item \exsen{Ddia midd de kälakälae dakädbaeneyo.} \extran{Little by little, they tried the deer meat.}
\end{itemize}
\item \anpl{kädkäd} \apl{kädnan} \defi{to remove bark, debark}
\begin{itemize}
\item \exsen{Ngäna put de näkädan.} \extran{I removed the bark.}
\end{itemize}
\item \anpl{käklläp} \apl{källpnan} \defi{to weed for the second time (easy work to remove the remaining plants), to sting, be painful, to be amused}
\begin{itemize}
\item \exsen{Bogo ttängäm de bäklläpän.} \extran{He will weed the garden.}
\item \exsen{Ngämo mokowang melem a klläpnan.} \extran{My favorite work is easy weeding.}
\end{itemize}
\item \anpl{käkllätt} \apl{kllättnan} \defi{to weed roughly (leaving bits behind), to fight, argue}
\begin{itemize}
\item \exsen{Ngäna tätäm towall de däkllätt.} \extran{Yesterday, I roughly weeded the grass.}
\item \exsen{Tätäm däkllättnegeya ngämo mäg alle.} \extran{Yesterday, my mother and I had an argument.}
\end{itemize}
\item \anpl{kälakäle} \apl{[]} \defi{to set (of the sun)}
\begin{itemize}
\item \exsen{Yäbäd a goklanän.} \extran{The sun was setting.}
\end{itemize}
\item \anpl{kälbae} \apl{kälbaenen} \defi{to singe (use brief heat to remove hair or down)}
\begin{itemize}
\item \exsen{Mälla da ttägäll baugän bottamänän, mätta de bakälbeaebneyo.} \extran{The woman will make mumu, finish, and then singe the yams.}
\end{itemize}
\item \anpl{kälkäl} \apl{kälmeny} \defi{to lie, fabricate}
\begin{itemize}
\item \exsen{Ngäna obo pallall kälkäl eran.} \extran{I'm making up stories about her.}
\end{itemize}
\item \anpl{källakälle} \apl{[]} \defi{to poison the river, to scrape food off the fire, e.g. banana, taro, yam}
\begin{itemize}
\item \exsen{walle kllaklle} \extran{to poison the river}
\end{itemize}
\item \anpl{källmakällme} \apl{källmanen} \defi{to survive}
\begin{itemize}
\item \exsen{Da ttongo kantri me lla da bompallängkmenynegän obaoba komlla kullum e, obaoba däbe kantri da mɨnyi ddone bakällmewän.} \extran{If the people in a country divide themselves into two groups, that country will not survive.}
\end{itemize}
\item \anpl{källttakälltte} \apl{källttanen} \defi{to cast away, expel}
\begin{itemize}
\item \exsen{Yesu gagäll anyke de däkällttawän lla bo guwo watt de.} \extran{Jesus cast the bad spirit out from the man's heart.}
\end{itemize}
\item \anpl{kälängkäl} \apl{[]} \defi{to climb, ascend}
\begin{itemize}
\item \exsen{Lla da llo de kälngkäl eran.} \extran{The man is climbing the tree.}
\end{itemize}
\item \anpl{käme näkäp me ddäganen} \apl{[]} \defi{to memorize}
\item \anpl{kängkäl} \apl{[]} \defi{to ascend, climb, go up, rise, to ascend, climb, go up}
\begin{itemize}
\item \exsen{Yäbäd a dingkälän tuk e.} \extran{The sun rose up into the sky.}
\item \exsen{Lla da llo de kälngkäl eran.} \extran{The man is climbing the tree.}
\item \exsen{Bogo manggo de dängkälän.} \extran{She climbed the mango tree.}
\end{itemize}
\item \anpl{kängkäm} \apl{kämnan} \defi{to squeeze, press}
\begin{itemize}
\item \exsen{Ge kup a gärep käp yid kämnan e dan.} \extran{This hole is for pressing grape juice (i.e. a winepress).}
\item \exsen{Ubi sana kämnanang dag.} \extran{They are squeezing sago.}
\end{itemize}
\item \anpl{kättkätt} \apl{kättnan} \defi{to fence, wall, build a fence or wall}
\begin{itemize}
\item \exsen{Ngäna polle de käg alle kättkätt eran.} \extran{I'm making a fence with the palm.}
\item \exsen{Bogo dädär alle polle de däkättän.} \extran{He built a wall out of stone.}
\end{itemize}
\item \anpl{kɨk} \apl{kɨnen} \defi{to crumple}
\item \anpl{kɨllakɨlle} \apl{kllanen} \defi{to scrape}
\item \anpl{kɨllkɨll} \apl{[]} \defi{to dig}
\begin{itemize}
\item \exsen{Auma däkällallo.} \extran{They dig a grave.}
\item \exsen{Gottamänän dibaballe kup di däkällneg a pos de däganeg.} \extran{I finished, and then I dug the holes and put the posts in.}
\end{itemize}
\item \anpl{laem} \apl{laemnen} \defi{to roll, wrap, to argue}
\begin{itemize}
\item \exsen{laemnenatt otät} \extran{rolled food, wrap}
\item \exsen{Bogo ttäkäll de dälemnegän.} \extran{She rolled up balls of yam and coconut.}
\end{itemize}
\item \anpl{liglig} \apl{lignen} \defi{to have sex, copulate}
\begin{itemize}
\item \exsen{Bogo liglig eran.} \extran{He is having sex.}
\item \exsen{Ubi do me alignallo.} \extran{They two had sex there.}
\item \exsen{Ubi iddob me ddone bulignegnän.} \extran{They will not be having sex at night.}
\end{itemize}
\item \anpl{llakek} \apl{llaknen} \defi{to collapse}
\begin{itemize}
\item \exsen{Särem a ollakan.} \extran{Darkness collapses.}
\end{itemize}
\item \anpl{llakllek} \apl{llaknen} \defi{to destroy}
\begin{itemize}
\item \exsen{Tätäm ttäle bun parga de bob a era dällekän.} \extran{Yesterday the flood destroyed the Ttäle Bun bridge.}
\end{itemize}
\item \anpl{llallem} \apl{llamnen} \defi{to finish weaving}
\item \anpl{llan gonggo} \apl{gonen} \defi{to turn one's ear, listen intently}
\begin{itemize}
\item \exsen{Ubi tätäm kaemne de llan donggoeyo.} \extran{Yesterday, they listened intently to the bee.}
\end{itemize}
\item \anpl{llanded} \apl{llandnen} \defi{to clear, to clarify, make clear, explain}
\begin{itemize}
\item \exsen{Oblle nyongo de dallendneyo.} \extran{They were clearing the road for her.}
\item \exsen{Bogo eka midd de dallendän.} \extran{He clarified the meaning.}
\end{itemize}
\item \anpl{llatat} \apl{[]} \defi{to trace, track, follow the blood (of an animal to kill it)}
\begin{itemize}
\item \exsen{Gongosän ma we, däräng a wa kumuddäga lla da dokomän, ddia llatat e.} \extran{He returned home, and took his dog and three men to track the deer.}
\end{itemize}
\item \anpl{llatat} \apl{llatnen} \defi{to twist, wrap}
\begin{itemize}
\item \exsen{Gull a kab llatnenatt alle iatt kollba gäddnan ma dan.} \extran{A net is a thing for catching fish, woven from twisted string.}
\end{itemize}
\item \anpl{llollo} \apl{[]} \defi{to pay respects}
\begin{itemize}
\item \exsen{Llabunda bi llollo e gobällaballe.} \extran{The relatives would go to pay their respects.}
\end{itemize}
\item \anpl{llollom} \apl{llomnen} \defi{to break, be damaged, to break}
\begin{itemize}
\item \exsen{Ekaklle we enanae gollomän kälakälae.} \extran{It hit the ground, smashing into pieces.}
\item \exsen{Bogo karpo inkätt de dullomän.} \extran{She broke the jar seal.}
\end{itemize}
\item \anpl{llullu} \apl{[]} \defi{to cut a tree so that one big tree falls onto many smaller trees and breaks them}
\item \anpl{lläblläb} \apl{lläbnan} \defi{to hide}
\item \anpl{llädae} \apl{llädaenen} \defi{to roll}
\begin{itemize}
\item \exsen{Llɨg a bal de llädaenen eran.} \extran{The boy is rolling the ball.}
\item \exsen{Ubi dädär ulle de dalldaeyo auma ud de däpakeyo.} \extran{They rolled a large rock and blocked the grave entrance.}
\end{itemize}
\item \anpl{llädäd} \apl{llädnan} \defi{to grab, get, catch, to buy, to marry, to come, arrive (figuratively)}
\begin{itemize}
\item \exsen{Ngäna walle we gäbän allan ddia llädäd e.} \extran{I am jumping in the water to grab the deer.}
\item \exsen{Ubi guspullän, be ubi obaoba gollädnegän.} \extran{They fell, but they caught themselves.}
\item \exsen{Iba pätt a ttongdae dan, ako kame ddone bällädeya.} \extran{We (incl.) only have one body; we don't get another.}
\item \exsen{Da ttongo lla da bäne baba bälle ten taosen mani bonttogän, ende bogo bällädän?} \extran{If someone gave your dad ten thousand kina, what would he buy?}
\item \exsen{Tu andred alle dällädeyo ttoe de.} \extran{He bought the skin for two hundred.}
\item \exsen{E mälla de ke bongo dälläd?} \extran{Which woman did you marry?}
\item \exsen{Da bongo lla tupi di nälläd, bongo mɨnyi llɨg kamebiag ag.} \extran{If you marry the tall man, you will have many children.}
\item \exsen{Angde obo täräp a ngäs e dällädalle, bogo mɨnyi do alle olle gogalle oba pate.} \extran{When his time to return came, he would call out to them.}
\end{itemize}
\item \anpl{lläklläk} \apl{[]} \defi{to spread fire}
\begin{itemize}
\item \exsen{Sali yu di mermerae lläklläk eran.} \extran{Sali is spreading the fire nicely.}
\end{itemize}
\item \anpl{llätät} \apl{llätnan} \defi{to get rid of oven stones}
\begin{itemize}
\item \exsen{Llɨg kälakäle, ttägäll käp de nälläteyoǃ} \extran{Small children, get rid of these mumu stones!}
\end{itemize}
\item \anpl{llɨtɨt} \apl{llɨtnen} \defi{to tell, report, say, to sing}
\begin{itemize}
\item \exsen{Bogo eka de llɨtɨtang meae nägnan.} \extran{He told the same story repeatedly.}
\item \exsen{Män a llɨg bälle tongoeang ttoen di dɨllɨtän.} \extran{The girl told the boy a funny story.}
\item \exsen{Ubi ddone amlle ende golläntmenynegän adawatta ddobaeddobae lel gognegnän.} \extran{They didn't say anything to anyone because they were very afraid.}
\item \exsen{Wagiba bandra de nɨllɨtnan sisri ag me.} \extran{Wagiba was singing this morning.}
\end{itemize}
\item \anpl{llɨtɨt} \apl{[]} \defi{to butcher, cut}
\begin{itemize}
\item \exsen{Ddia ddäg de nällɨt.} \extran{[You] cut the deer back.}
\end{itemize}
\item \anpl{lugoe} \apl{lugoenen} \defi{to drag}
\item \anpl{lulu} \apl{[]} \defi{to gossip about, tell rumors about}
\begin{itemize}
\item \exsen{Luluang lla da ai dan gagäll ud de bätramän.} \extran{One who gossips can open the door to bad things.}
\item \exsen{Ngäna dandärmällnegne mälla de ami bam daluneyo.} \extran{I heard the women who were gossiping about you.}
\end{itemize}
\item \anpl{lɨklɨk} \apl{[]} \defi{to melt}
\begin{itemize}
\item \exsen{Ngämo ine kutt a daeya yu da lɨklɨk eran.} \extran{Fire is melting my water container.}
\end{itemize}
\item \anpl{malmal} \apl{malnen} \defi{to mark land (for planting or settlement), to accuse, blame}
\begin{itemize}
\item \exsen{Bobäll ttängäm malmal e.} \extran{Let's go mark a garden spot.}
\item \exsen{Ge ma de dämaleyo.} \extran{They chose to settle this area.}
\item \exsen{Ubi yuwog abal ttoen e bam malmal nallo.} \extran{They are accusing you of so many things.}
\end{itemize}
\item \anpl{malmel} \apl{[]} \defi{to squeeze (e.g. a hand)}
\item \anpl{mamoe} \apl{maenen} \defi{to hunt, go hunting, hunt}
\begin{itemize}
\item \exsen{Obo moko da mamoe e dan.} \extran{He wants to go hunting.}
\item \exsen{Ubi därängmeny damoenegnän.} \extran{They were hunting without dogs.}
\item \exsen{Namoeaemnalla.} \extran{We were hunting them.}
\item \exsen{Mamoe a angde gongkamalle, ddob lla da mɨnyi po we ngattong e gobällalle.} \extran{When the hunt starts, some men will go first to block the way.}
\end{itemize}
\item \anpl{mattmett} \apl{[]} \defi{to put in oven}
\begin{itemize}
\item \exsen{Sɨmell gullbe de ddob sana peyang dazgiaebeya, ada dämetteyo ttägäll ik i.} \extran{We wrapped the boar with some sago and put it in the mumu.}
\end{itemize}
\item \anpl{mermer} \apl{mernan} \defi{to shrink, shrivel}
\begin{itemize}
\item \exsen{Ge mätta da mermer allan.} \extran{This yam is shrinking.}
\end{itemize}
\item \anpl{mikmik} \apl{[]} \defi{to suck}
\item \anpl{mirmir} \apl{mirnen} \defi{to lick}
\item \anpl{modgae} \apl{[]} \defi{to chase away, potentially an Agob word?}
\item \anpl{mokmok} \apl{moknen} \defi{to wipe; dry by wiping}
\item \anpl{mäkamäke} \apl{mäkanen} \defi{to use}
\begin{itemize}
\item \exsen{Ibi ge up de mullamulla we mäkanen eralla.} \extran{We use this banana for medicine.}
\end{itemize}
\item \anpl{mälamäle} \apl{mälanen} \defi{to patch, to dress (a wound)}
\begin{itemize}
\item \exsen{Känaebag, ngäna mɨnyi ute de bamle.} \extran{Tomorrow, I will dress the wound.}
\end{itemize}
\item \anpl{mällamälla} \apl{mällanen} \defi{to tie}
\begin{itemize}
\item \exsen{Yu wi daspuniyu, mällamällawatt de, ada oba yu a bättämän.} \extran{They threw him on the fire and tied him so that the fire would burn him.}
\item \exsen{Ttongo lla da ddone mullae dan mängallang lla bo ma we bozenän obo za gämäll e, be dam de da bogo ngattong däbe lla de bämllawän.} \extran{A person cannot enter the house of a powerful man to steal his things unless they tie that man up first.}
\end{itemize}
\item \anpl{mällkae} \apl{[]} \defi{to bend}
\begin{itemize}
\item \exsen{Kote dämällkaenän obo pate.} \extran{He was bowing his head (lit. bent nape) to her.}
\end{itemize}
\item \anpl{mälmäl} \apl{mälnan} \defi{to squeeze, accuse}
\item \anpl{mänddmändd} \apl{mänddnan} \defi{to drown, struggle in water, to drown, force underwater, sink}
\begin{itemize}
\item \exsen{Bogo mänddmändd e gongkamän walle we.} \extran{She started to drown in the water.}
\item \exsen{Bogo aoli gomänddän walle me.} \extran{She almost drowned in the water.}
\item \exsen{Bogo ddia de ine dämänddän.} \extran{He drowned the deer in the water.}
\end{itemize}
\item \anpl{mändmänd} \apl{mändnan} \defi{to raise, rear, to feed, help}
\begin{itemize}
\item \exsen{Ubi obom damändeyo.} \extran{They raised him.}
\item \exsen{Indonesia me gänya ddäddäg de damändaebneyo.} \extran{In Indonesia, they tamed this animal.}
\item \exsen{Ngäna obom kumuddäga nge alle bamänd.} \extran{I will feed him three coconuts.}
\end{itemize}
\item \anpl{mänymäny} \apl{[]} \defi{to vomit}
\item \anpl{mättae} \apl{mättaenen} \defi{to threaten, raise fists}
\begin{itemize}
\item \exsen{Ngämo mälla bom mättaenen nängkaman.} \extran{I started to threaten my wife.}
\item \exsen{Ngäna obom damttaene.} \extran{I was threatening him.}
\end{itemize}
\item \anpl{mättmätt} \apl{mättnan} \defi{to wear, dress oneself, put on, to dress}
\begin{itemize}
\item \exsen{Bogo banggu mättmätt allan.} \extran{He is wearing a headdress.}
\item \exsen{Mägda sisor kaptte amättan.} \extran{Mother put on new clothes.}
\item \exsen{Ngäna obom mättmätt eran.} \extran{I am dressing her.}
\end{itemize}
\item \anpl{nane} \apl{nanen} \defi{to drink}
\begin{itemize}
\item \exsen{Bogo nane eran.} \extran{He is drinking.}
\item \exsen{Ngäna konkonang ine de bäna.} \extran{I will drink the beer.}
\item \exsen{Ngämi dänaeyaemnalla.} \extran{We (excl.) were drinking.}
\end{itemize}
\item \anpl{nganae} \apl{nganaenen} \defi{to coil, go around, to spin, rotate}
\begin{itemize}
\item \exsen{Mätta da dade de nganae eran.} \extran{The yam is coiling around the yam stick.}
\item \exsen{Ekaklle da nganaenen allan.} \extran{The earth is spinning.}
\end{itemize}
\item \anpl{ngange} \apl{nganen} \defi{to communicate, deliver a message}
\begin{itemize}
\item \exsen{ngangema ttoen} \extran{method of communication}
\item \exsen{Lla da mɨnyi utt alle eka de bängaeyo.} \extran{Men will deliver the message with the conch shell.}
\end{itemize}
\item \anpl{ngangem} \apl{[]} \defi{to adopt}
\begin{itemize}
\item \exsen{Ngäna ge män de obaene de dangem.} \extran{I adopted this girl from them.}
\end{itemize}
\item \anpl{ngangleb} \apl{ngalbänen} \defi{to look for, search for}
\begin{itemize}
\item \exsen{Lla da mälla de dängalbänän.} \extran{The man searched for the woman.}
\end{itemize}
\item \anpl{ngasnges} \apl{ngasnen} \defi{to do, to make, to happen}
\begin{itemize}
\item \exsen{Ende bongo ngasnges eralle ttongo lla bo pate, bongo ngasnges eralle God bo pate.} \extran{What you do to another person, you do to God.}
\item \exsen{Bongo bäne kumuddäga melem de nangesneg.} \extran{Do your three chores.}
\item \exsen{Ibi ibra} \extran{We (excl.) will make a new fireplace for us.}
\item \exsen{Masar ine kube de ngämo nangesan.} \extran{Grandfather made my water bucket.}
\item \exsen{Ge ttoen a gongesän ngämo patme.} \extran{This thing happened to me.}
\end{itemize}
\item \anpl{ngatae} \apl{ngataenen} \defi{to visit}
\begin{itemize}
\item \exsen{Ngämi mɨnyi ngataenen e buddällmam.} \extran{We (excl.) will come to visit.}
\end{itemize}
\item \anpl{ngetae} \apl{[]} \defi{to touch, to get engaged, to unite, join, to grip, hold onto}
\begin{itemize}
\item \exsen{Ttongo lla da mälla da walle nägabällallo ddɨg alle angeteallo.} \extran{A man was standing with his wife with their backs touching.}
\item \exsen{Lla da medäda wa mägda de wanseg yaran a obo mälla peyang ngetae allan.} \extran{The man leaves his father and mother and gets engaged to his woman.}
\item \exsen{Lla da ddone mɨnyi sapasapang beyagän Adi bäne enanae ngetaeatt de.} \extran{Man shall not separate what God has joined together for good.}
\item \exsen{Ngäma ttang de deyangeteyo.} \extran{They gripped our (excl.) hands.}
\end{itemize}
\item \anpl{ngongo} \apl{ngonen} \defi{to smooth, sand (a surface)}
\begin{itemize}
\item \exsen{Turik de nongowan.} \extran{I smoothed the axe.}
\end{itemize}
\item \anpl{ngongop} \apl{[]} \defi{to hug, embrace}
\begin{itemize}
\item \exsen{Mäg da llɨg kälsre de nongopan.} \extran{The mother hugged her little boy.}
\end{itemize}
\item \anpl{ngonoe} \apl{ngonoenen} \defi{to ask}
\begin{itemize}
\item \exsen{Llamda da abo ubim ngonoenen de deyangkamän.} \extran{The old man started to question those two.}
\item \exsen{Ubi obaoba gongnoemenynegnän ada, "Ge endan?"} \extran{They were asking themselves, "What is this?"}
\end{itemize}
\item \anpl{ngädngäd} \apl{ngädnan} \defi{to roll up, curl, to roll up, curl, to fold}
\begin{itemize}
\item \exsen{Gullem a gongädän.} \extran{The snake coiled up.}
\item \exsen{Ngämi tater nängädaeballa.} \extran{We (excl.) rolled up the mats.}
\item \exsen{Bogo ttang nängädnegan.} \extran{She folded her arms.}
\end{itemize}
\item \anpl{ngällae} \apl{[]} \defi{to look back}
\begin{itemize}
\item \exsen{Ngäna ngllae allan.} \extran{I am looking back.}
\item \exsen{Bogo mäse dinduag me ako gongllaeyän däräng bälle, be ddone obom ikop dägagän.} \extran{While running, she tried to look back for the dog, but she couldn't see him.}
\end{itemize}
\item \anpl{ngällngäll} \apl{ngällnan} \defi{to produce, yield, bear (fruit)}
\begin{itemize}
\item \exsen{Wit däm a ulleulle gognegän a käp de yuwog gongällanegän.} \extran{The wheat grew big and produced lots of grain.}
\item \exsen{Nge da ngällngällang dan.} \extran{The coconut tree is bearing fruit.}
\end{itemize}
\item \anpl{ngämae} \apl{[]} \defi{to go around, take the long way}
\begin{itemize}
\item \exsen{Kurupel gungmaeyän ttongo nyongo dae dallän do auma.} \extran{Kurupel went using another path to the grave.}
\end{itemize}
\item \anpl{ngänngän} \apl{ngännan} \defi{to swear}
\item \anpl{ngänyngäny} \apl{ngänynan} \defi{to swallow}
\begin{itemize}
\item \exsen{Amtet de nängänymällneg!} \extran{Take (lit. swallow) a breath!}
\end{itemize}
\item \anpl{ngätae} \apl{ngätaenen} \defi{to check}
\begin{itemize}
\item \exsen{Lla da mɨnyi ngätaenen e ballnän.} \extran{The man will be going to check.}
\end{itemize}
\item \anpl{ngättangätta} \apl{ngättanen} \defi{to block, obscure, to occupy, take over}
\begin{itemize}
\item \exsen{Bongo ddone ikop dagne, llo da ngänäm dangttawalle.} \extran{You couldn't see me; the tree was obscuring me.}
\item \exsen{Särem a dangttanän ttängäm de.} \extran{Darkness took over the village.}
\end{itemize}
\item \anpl{ngättangätte} \apl{ngättanen} \defi{to count, to read}
\begin{itemize}
\item \exsen{Ingglis eka walle dangättnealle.} \extran{I was counting in English.}
\end{itemize}
\item \anpl{nyongkoe} \apl{nyongkoenen} \defi{to pull}
\begin{itemize}
\item \exsen{Angde tudi gognän, obo tudi di kollba ulle da danykoeyän.} \extran{While fishing, a big fish pulled on her line.}
\end{itemize}
\item \anpl{nyägae} \apl{nyägaenen} \defi{to flail, to stir}
\begin{itemize}
\item \exsen{Guzi da nyägaenen allan.} \extran{The crayfish was flailing around.}
\item \exsen{Mälla da wätät de nanyägaenen spun alle.} \extran{The woman was stirring the food with a spoon.}
\item \exsen{Danygaeaebneyo sana peyang.} \extran{They were stirring them together with sago.}
\end{itemize}
\item \anpl{nyämaenyämae} \apl{[]} \defi{to surround}
\begin{itemize}
\item \exsen{Lla da obom danymaeyo.} \extran{People surrounded him.}
\end{itemize}
\item \anpl{nyäng} \apl{[]} \defi{to dance}
\begin{itemize}
\item \exsen{Dänyängnän obo peyang.} \extran{They were dancing with him.}
\end{itemize}
\item \anpl{nyängoe} \apl{nyängoenen} \defi{to walk slowly with someone whose leg is hurt}
\begin{itemize}
\item \exsen{Danyängoeneyo gänye ma dälltam.} \extran{We walked slowly and came up to the house.}
\end{itemize}
\item \anpl{nyänye} \apl{nyänan} \defi{to share, split}
\begin{itemize}
\item \exsen{Ngämi dänyeya däbe ttägäll käp de.} \extran{We (excl.) shared that money.}
\item \exsen{Otät de nyänan amallo.} \extran{They're sharing the food.}
\end{itemize}
\item \anpl{nyäroe} \apl{nyäroenen} \defi{to creep}
\begin{itemize}
\item \exsen{Angde ngäna ikop däga ada ddia dan, ngäna nyäroe de dängkam.} \extran{When I saw the deer, I started to creep up.}
\end{itemize}
\item \anpl{nyärpae} \apl{[]} \defi{to refuse (someone in the dative)}
\begin{itemize}
\item \exsen{Däbe män a oblle nyärpae ma da ddone mulldae gogon.} \extran{That girl couldn't refuse him.}
\end{itemize}
\item \anpl{näkäp ttomoe} \apl{ttomoenen} \defi{to worry}
\begin{itemize}
\item \exsen{Bogo näkäp ttomoenen me giddollnen allan.} \extran{He is living worrying.}
\end{itemize}
\item \anpl{ogo} \apl{[]} \defi{hoard}
\item \anpl{okok} \apl{oknen} \defi{to rub}
\item \anpl{omom} \apl{omnen} \defi{to sweep}
\begin{itemize}
\item \exsen{Ngäna ag me ma de noman.} \extran{I swept the house in the morning.}
\end{itemize}
\item \anpl{opap} \apl{opanen} \defi{to cross over, pass over, move across}
\begin{itemize}
\item \exsen{Ibi bopäll bem de apte menae e daupeya.} \extran{Let's cross the sea to the other side.}
\end{itemize}
\item \anpl{opop} \apl{[]} \defi{tangle, wind}
\item \anpl{otät} \apl{wätnan} \defi{to eat, food, hunger}
\begin{itemize}
\item \exsen{Diba toto we, bogo sana dorko de doton.} \extran{That evening, she ate dry sago.}
\item \exsen{Ubi llo käp otnanang dag.} \extran{They eat tree fruit.}
\item \exsen{Bogo mer wätät de yu dägagän.} \extran{She cooked nice food.}
\item \exsen{Otät atta ngämo kutt a mängallmeny agmallo.} \extran{My bones became weak from hunger.}
\end{itemize}
\item \anpl{pampem} \apl{pamnen} \defi{to fish}
\begin{itemize}
\item \exsen{Ag me källäm pamnen e gobäll.} \extran{In the morning, we went to fish at the pond.}
\end{itemize}
\item \anpl{panypeny} \apl{panynen} \defi{to speak, talk, say}
\begin{itemize}
\item \exsen{Ngäna Ende eka de panypeny eran.} \extran{I speak Ende.}
\item \exsen{Ngäna Ingglis eka, Motu eka de ddone bäpanyneg.} \extran{I can't speak English or Hiri Motu.}
\item \exsen{Bogo ge eka de däpanyän.} \extran{She said this.}
\end{itemize}
\item \anpl{paopao} \apl{[]} \defi{to cut around, prune (a plant)}
\begin{itemize}
\item \exsen{Däbe kapang ulle de näpao!} \extran{[You] prune the big acacia tree!}
\end{itemize}
\item \anpl{pape} \apl{panen} \defi{to cut (grass), to smash, crush (something soft, but not a flower)}
\begin{itemize}
\item \exsen{Kitar pu de däpeyo.} \extran{They cut the kitar grass.}
\item \exsen{Ngäna mameat de näpewan.} \extran{I smashed the papaya.}
\item \exsen{Ubi mätta de bäpeaemneyo.} \extran{They will be smashing yams.}
\end{itemize}
\item \anpl{papllek} \apl{[]} \defi{to chop}
\begin{itemize}
\item \exsen{Bogo yu di däpellknegän.} \extran{She chops wood.}
\end{itemize}
\item \anpl{paplläg} \apl{pllagnen} \defi{to fly}
\begin{itemize}
\item \exsen{Ngämi pällgaeb amalla.} \extran{We are flying.}
\item \exsen{Pa da dapllägän matu we.} \extran{The bird flew down.}
\end{itemize}
\item \anpl{papoe} \apl{[]} \defi{to pierce, make a hole}
\begin{itemize}
\item \exsen{Bogo obo llan de napoeyan.} \extran{She pierced her ear.}
\item \exsen{Ngäna nyäng tär ngasnges e nyäng de napoenan waeya käsre alle.} \extran{I used the small wire to make holes in the bag to put the new bag string.}
\end{itemize}
\item \anpl{papälläk} \apl{pallknen} \defi{to split, chop}
\begin{itemize}
\item \exsen{Bogo yu di däpellknegän.} \extran{She split the wood.}
\end{itemize}
\item \anpl{pedae} \apl{pedaenen} \defi{to sweep}
\begin{itemize}
\item \exsen{Ma de dapedealle.} \extran{He sweeped the house.}
\end{itemize}
\item \anpl{pentae} \apl{[]} \defi{to spread, be transmitted, to transfer, transmit, spread, pass on, pass down}
\begin{itemize}
\item \exsen{Ge itrell a mɨnyi bompetaeyän.} \extran{This disease will spread.}
\item \exsen{Ge itrell a mängalae lla yaba pate pentaeang dan.} \extran{This disease spreads to others quickly.}
\item \exsen{Ende lla da oba omog ttoen tameny de oba llɨg aba pate dampetaeneyo.} \extran{Ende people were passing their sorcery on to their children.}
\end{itemize}
\item \anpl{pipi} \apl{pinen} \defi{to shoot, spear}
\begin{itemize}
\item \exsen{Ngäna ddob mozaya kollba de däpineg.} \extran{I speared some mozaya fish.}
\end{itemize}
\item \anpl{pipllug} \apl{pillugnen} \defi{to fly}
\begin{itemize}
\item \exsen{Bongo mɨnyi ikop nägneg ddob apapi di bipllugaebnän.} \extran{You will see some butterflies flying.}
\end{itemize}
\item \anpl{piro ttängattänge} \apl{[]} \defi{to become unconscious}
\begin{itemize}
\item \exsen{Bogo piro nangttängenan.} \extran{He fell down unconcsious.}
\end{itemize}
\item \anpl{pitkae} \apl{pitkaenen} \defi{to untie}
\begin{itemize}
\item \exsen{Ngäna mɨnyi bomällkam a obo nying kollop tär de bäpitkaeneg.} \extran{I will bend down and untie his sandal laces.}
\end{itemize}
\item \anpl{pittpitt} \apl{pittnen} \defi{to weave, to sew, stitch}
\begin{itemize}
\item \exsen{Sana dor de adingoll däpittaemeya.} \extran{We weaved the sago stalks like this.}
\item \exsen{Amo umllang dan pittpitt a?} \extran{Who here knows how to sew?}
\item \exsen{Pittnen de dängkaemnegän.} \extran{She started to stitch (the wounds).}
\end{itemize}
\item \anpl{po} \apl{ponen} \defi{to pierce, to split}
\begin{itemize}
\item \exsen{Mälläng e dapoeaemeyo.} \extran{They pierced noses.}
\end{itemize}
\item \anpl{pollpoll} \apl{pollnen} \defi{to bark (at)}
\begin{itemize}
\item \exsen{Ge däräng a sämell ulle de dupollneyo.} \extran{These dogs were barking at the big pig.}
\item \exsen{Kwalde bogo ada gongnomenynän ada ka däräng a pollnenang me erallo däbe sämell de.} \extran{Kwalde thought that the dogs were still barking at the pig.}
\end{itemize}
\item \anpl{popllem} \apl{[]} \defi{to flap}
\begin{itemize}
\item \exsen{Kaptte da wel me angde popllem anggan, dedam era mamaldae pätnan anggan.} \extran{When the clothes flap in the wind, they dry quickly.}
\end{itemize}
\item \anpl{popllo} \apl{[]} \defi{twist}
\item \anpl{popo} \apl{ponen} \defi{to sharpen, to carve}
\begin{itemize}
\item \exsen{Dompa de däpoaemneyo.} \extran{They were sharpening the arrows.}
\item \exsen{Ge pensil a popoatt dan.} \extran{This pencil is sharpened.}
\item \exsen{Gall a llo popoatt dan.} \extran{The canoe is carved wood.}
\item \exsen{Bägäl de popo eran.} \extran{He shapes the bow.}
\end{itemize}
\item \anpl{potpot} \apl{[]} \defi{to slice open}
\begin{itemize}
\item \exsen{Bogo sɨmell käm dupotän.} \extran{He sliced open the pig stomach.}
\end{itemize}
\item \anpl{päddpädd} \apl{päddnen} \defi{to grow, to explode}
\begin{itemize}
\item \exsen{ine toko me päddnenang towall} \extran{grass growing on the water's surface}
\item \exsen{Däm kutt a ekaklle me guspullän a däpäddän.} \extran{The seed fell to the ground and grew.}
\item \exsen{Yu me pattlle da päddpädd allan.} \extran{The bamboo explodes in the fire.}
\end{itemize}
\item \anpl{pädoe} \apl{pädoenen} \defi{to blow, to blow}
\begin{itemize}
\item \exsen{Wel a gopädoeän.} \extran{The wind blew.}
\item \exsen{Mälla da wa lla da utt de dapädoenegeyo.} \extran{The man and women blew conch shells.}
\end{itemize}
\item \anpl{pädrall} \apl{[]} \defi{to spread (out), scatter, disperse, to spread (out), scatter, sow}
\begin{itemize}
\item \exsen{Lla da oba mällamällong gupädrellän otät ma.} \extran{The men and their wives spread out to get food.}
\item \exsen{Oba näkäpngon a tämamae apädrellan.} \extran{Their thoughts are all scattered.}
\item \exsen{däm kutt pädrallag lla} \extran{man sowing seeds}
\item \exsen{Bogo up di dotnegnän a ttoe dapädrällnegnän.} \extran{He was eating bananas and scattering the peels.}
\end{itemize}
\item \anpl{pällganen} \apl{[]} \defi{to hang}
\begin{itemize}
\item \exsen{Ttongo mälla da iddpo de nällpäganegnan.} \extran{A woman was hanging the clothes.}
\end{itemize}
\item \anpl{pämpllätt} \apl{plläntnan} \defi{to become healthy; grow}
\item \anpl{pänae} \apl{pänaenen} \defi{to turn back, turn around, to capsize, turn over, flip over, to turn, become, transform, to translate, zenith (position of the sun at high noon)}
\begin{itemize}
\item \exsen{Ubi komlla gopnaeyo a ngäsengäse dindugiyu.} \extran{The two of them turned around and ran the other way.}
\item \exsen{Gopnaeyän.} \extran{It capsized.}
\item \exsen{Ngämaene gall a ada ka bopänayän, adawatta däbe za da ngämim ge deyangkollmällnän.} \extran{Our (excl.) boat nearly capsized because that thing was following us.}
\item \exsen{Obo mikutt a ttänttämang e gopnaeyän.} \extran{His anger made him hot.}
\item \exsen{Lla daeya Moli, be angde walle we guspunän, bunkuttang e gopnaen.} \extran{Moli was a man, but when he jumped into the water, he turned into a catfish.}
\item \exsen{Ngäna ttoenttoen de näpänaeyan Ende we.} \extran{I translated this story into Ende.}
\end{itemize}
\item \anpl{pänyae} \apl{pänyaenen} \defi{to hop}
\begin{itemize}
\item \exsen{Bogo pänyaemällang dan.} \extran{He hops slowly.}
\end{itemize}
\item \anpl{päp} \apl{pänan} \defi{to draw or mark, to speed up}
\item \anpl{pätpät} \apl{pätnan} \defi{to dry}
\item \anpl{pättapätte} \apl{pättanen} \defi{to shake off, to hit grass with a stick to scare away animals}
\item \anpl{pättol} \apl{pättolnen} \defi{to start singing}
\begin{itemize}
\item \exsen{Bogo bandra de dipttolän.} \extran{He starting singing a song.}
\end{itemize}
\item \anpl{pɨnypɨny} \apl{pɨnynen} \defi{to plant a lot, many stars appearing}
\begin{itemize}
\item \exsen{Piro da pɨnypɨny eran.} \extran{Many stars are coming out.}
\end{itemize}
\item \anpl{rorwae} \apl{[]} \defi{to stagger (e.g. when drunk)}
\begin{itemize}
\item \exsen{Rorwae a mudan, llokttangae nägabäll.} \extran{Don't stagger; stand straight.}
\end{itemize}
\item \anpl{rullgoe} \apl{[]} \defi{to drag}
\begin{itemize}
\item \exsen{Tim ngänäm dangminggän ngämi däbe käza de tutu wi darullgoeya.} \extran{Tim helped me drag that crocodile onto land.}
\end{itemize}
\item \anpl{sae} \apl{saenen} \defi{to close, cover, to extinguish, put out}
\begin{itemize}
\item \exsen{Nga bongo ikop nasǃ} \extran{You, close your eyes!}
\item \exsen{Däräng a ikop dasän.} \extran{The dog closed his eyes.}
\item \exsen{Yu da saemeny dan.} \extran{The fire isn't put out.}
\end{itemize}
\item \anpl{soroe} \apl{[]} \defi{to try, attempt, to challenge, try, test; tempt, test, exam}
\begin{itemize}
\item \exsen{Ongg kame ako dasäroeyän.} \extran{Ongg tried it again.}
\item \exsen{Ubi mɨnyi oba mängall alle bosroenegnän ngämene eka ngänaem e.} \extran{They will be trying hard to understand my words.}
\item \exsen{Ubi ngonoe eka walle obom dasroeyo.} \extran{They challenged him with a question.}
\item \exsen{Bogo dedme poti ebdo dagirnän a saeten dasroenän.} \extran{He stayed there for forty days and Satan was tempting him.}
\item \exsen{Amo bun da soroe me mullae gogalle, bogo mɨnyi dallalle tuk skul e.} \extran{He who does well on the test will go to secondary school.}
\end{itemize}
\item \anpl{sorsor} \apl{[]} \defi{to affect at a distance. For example, if poison is poured up stream, the fish down stream will also be affected.}
\item \anpl{sänasäne} \apl{[]} \defi{to take out}
\begin{itemize}
\item \exsen{Ngämi ddäddäg de mɨnyi bäsnaemeyo säspen att.} \extran{We (excl.) will take out the meat from the pot.}
\end{itemize}
\item \anpl{särämbae} \apl{särämbaenen} \defi{to prepare, be prepared, be ready, get ready, to prepare, arrange, get ready, to fix, solve, resolve}
\begin{itemize}
\item \exsen{Tämamae ansärbemom!} \extran{Everyone, get ready!}
\item \exsen{Ngäna ada ngonongg allan ada malla bongo särämbaeatt dan.} \extran{I think that you are not prepared.}
\item \exsen{Ubi oba gall me dagwaeya, gull de dansärämbenegneyo.} \extran{The two of them were in their canoe, preparing their nets.}
\item \exsen{Ngäna obo pätt de nansärämbeyan au we.} \extran{I prepared his body for burial.}
\item \exsen{Ngämlle mokowang melem dan lla bo buddo särämbae.} \extran{I enjoy the task of solving people's problems.}
\end{itemize}
\item \anpl{säsäs} \apl{säsnan} \defi{to rub}
\begin{itemize}
\item \exsen{Gosäsneyo obaoba.} \extran{They were rubbing each other.}
\end{itemize}
\item \anpl{sɨs} \apl{sɨnen} \defi{to extinguish, turn off}
\begin{itemize}
\item \exsen{Ngäna tos de dɨs.} \extran{I turned off the light.}
\end{itemize}
\item \anpl{tanter} \apl{[]} \defi{to guard in one's sleep, sleep with}
\begin{itemize}
\item \exsen{Bogo gotarnän ma da däbe yunu danteralle.} \extran{She slept over, guarding that house.}
\end{itemize}
\item \anpl{tanträm} \apl{tarmnen} \defi{to smash}
\item \anpl{tongoe} \apl{tongoenen} \defi{to play, to laugh (at), game; sport}
\begin{itemize}
\item \exsen{Nga ubi dag tongoenen meae anggan.} \extran{They are still playing.}
\item \exsen{Llɨg kälekäle da ebdo ulle atongoenegnan.} \extran{The children were playing all day.}
\item \exsen{Ibi togotogol botongoe nalla.} \extran{Let's play hide and seek.}
\item \exsen{Ubi ttongo lla de tongoenen erallo.} \extran{They laugh at someone.}
\item \exsen{Lama käsre ngämira llɨtɨt dängkamän tongoetongoeang.} \extran{Lama started to tell us (excl.) the story, laughing.}
\item \exsen{Ngämo mokowang tongoe a basketbol dan.} \extran{My favorite game is basketball.}
\end{itemize}
\item \anpl{totoe} \apl{[]} \defi{conceive}
\item \anpl{totrop} \apl{tärpnen} \defi{to remove scales, to clean up, to scratch, like to scratch the tree with your fingers, to scrape, e.g. when you shoot a deer and the arrow just scratches the animal, to use the hoe or spade to remove small pieces of grass from a clear place}
\item \anpl{tratre} \apl{[]} \defi{hollow}
\item \anpl{trongoe} \apl{[]} \defi{to check a bucket or container for water}
\item \anpl{ttanttem} \apl{ttamnen} \defi{to be confused}
\begin{itemize}
\item \exsen{Obo eka me ubi gonttemnegän.} \extran{They were confused by his story.}
\end{itemize}
\item \anpl{ttatta} \apl{ttanen} \defi{to chop a tree}
\begin{itemize}
\item \exsen{Bogo llo de ttanen anggan.} \extran{He's chopping a tree.}
\end{itemize}
\item \anpl{ttattlläb} \apl{[]} \defi{to open}
\begin{itemize}
\item \exsen{Wäd a wättallban.} \extran{The door opened.}
\end{itemize}
\item \anpl{ttokoe} \apl{[]} \defi{to cut}
\begin{itemize}
\item \exsen{Dibaballe ngäna pos de ttokoenen de dängkamneg, tämamae naen dattkoeneg.} \extran{Afterwards, I started to cut the posts, I chopped all nine.}
\end{itemize}
\item \anpl{ttotto} \apl{ttonen} \defi{to tie, to collect}
\begin{itemize}
\item \exsen{Ngäna ngämo bägäl de ttotto dägneg toboll peyang.} \extran{I tied my bow and arrows together.}
\item \exsen{Bogo ekaklle me gontmonän up gälbänanatt ttonen e.} \extran{He waited on the ground to collect the fallen bananas.}
\end{itemize}
\item \anpl{ttäbattäbe} \apl{[]} \defi{to block}
\begin{itemize}
\item \exsen{Ttäbanen eran, ede ge nyongo da gagäll allan.} \extran{It's blocked, so this road is no good.}
\end{itemize}
\item \anpl{ttäkattäke} \apl{[]} \defi{to fold}
\item \anpl{ttäkoe} \apl{ttäkoenen} \defi{to chop, cut down, mow; shave}
\begin{itemize}
\item \exsen{Ngäna pos de ttokoenen de dängkamneg, tämamae naen dattkoeneg.} \extran{I started to cut the posts; I chopped all nine.}
\item \exsen{Dade towall de battkoe.} \extran{Maybe I'll mow the grass.}
\item \exsen{Ngäna obo ttatt kom de nattkoeyan.} \extran{I shaved his beard.}
\end{itemize}
\item \anpl{ttämattäme} \apl{ttämanen} \defi{to be crowded}
\begin{itemize}
\item \exsen{Ngättma da lla walle gonttämawän.} \extran{The place was crowded with people.}
\end{itemize}
\item \anpl{ttängattänge} \apl{ttänganen} \defi{to read, recite, date}
\begin{itemize}
\item \exsen{Ddob llaeyaba eka walle darbnen att eka da, ngäna mɨnyi bangttängeneg.} \extran{Stories written in other people's languages, I will read them.}
\item \exsen{Sisri ngasekäma ttängattänge da endan?} \extran{What's the date today?}
\end{itemize}
\item \anpl{ttätt} \apl{[]} \defi{to urinate, surprise}
\item \anpl{ttättälläg} \apl{ttällägnen} \defi{to tear, tear up}
\begin{itemize}
\item \exsen{Gudne iddpo ttättllägatt pop de mäzi abal ulle ubemang bägagän.} \extran{It will make the torn holes in the old clothing even wider.}
\end{itemize}
\item \anpl{tulgoe} \apl{tulgoenen} \defi{to gossip about}
\begin{itemize}
\item \exsen{Tätäm mälla da ngänäm era daitulgoeneyo duli.} \extran{Yesterday, women were gossiping about me there.}
\end{itemize}
\item \anpl{tunen} \apl{[]} \defi{to bump, obligatory plural object (even if singular object)}
\item \anpl{tutu} \apl{tunen} \defi{to knock fruit continuously}
\item \anpl{täbatäbe} \apl{täbanen} \defi{to plan, plan}
\begin{itemize}
\item \exsen{Ngäna gotäba ngämo nge ibeny e.} \extran{I planned to plant my coconut.}
\item \exsen{Ngämi diba toto me ngäma täbatäbe de dangeseya iddob me ddia ma koenmäll e.} \extran{We (excl.) made the plan that evening to chase the deer at night.}
\end{itemize}
\item \anpl{täli} \apl{tälinin} \defi{to repeat}
\begin{itemize}
\item \exsen{Bogo dallän ikopse we, dantlinigän däbem eka dae.} \extran{He went to pray, repeating just those words.}
\end{itemize}
\item \anpl{täratäre} \apl{täranen} \defi{to dig out, hollow out}
\item \anpl{tärnetärne} \apl{[]} \defi{to peck}
\begin{itemize}
\item \exsen{tärnetärneang dan} \extran{he is pecking}
\end{itemize}
\item \anpl{täträk} \apl{tärknan} \defi{to go underneath, to push in, push through}
\begin{itemize}
\item \exsen{Käsre dibaballe ada gogon goträkän.} \extran{Then from there, he went underneath.}
\item \exsen{Ngämo giri de käza bo ume ttäp e däträkne.} \extran{I was pushing in my knife into the crocodile's mouth.}
\end{itemize}
\item \anpl{tätäräp} \apl{täräpnan, tarpnen} \defi{to cut, to cut across, take a shortcut, man-made clearing}
\begin{itemize}
\item \exsen{Ibi goeg tatäräp e.} \extran{Let's go to the garden to cut down trees.}
\item \exsen{Ngäna ngämo llan de yatärpänan.} \extran{I cut my ears.}
\item \exsen{Minong gänyaollemae ditäräpmällnän ngämo pate.} \extran{Minong took a shortcut this way towards me.}
\end{itemize}
\item \anpl{tɨt} \apl{tɨnen} \defi{to beat sago, pound sago, to hollow out, dig out}
\begin{itemize}
\item \exsen{Ubi sana pätt att de tɨt erallo a kängkäm erallo.} \extran{They pound the pith of the sago palm and squeeze it.}
\item \exsen{Bogo mɨnyi sana bɨtɨn.} \extran{He will pound sago.}
\item \exsen{Ngäna gall de tɨt iran.} \extran{I am digging out a canoe.}
\item \exsen{Bäne turik de abo iweny gall tɨt i.} \extran{Bring your axe to dig the canoe.}
\end{itemize}
\item \anpl{tɨtɨp} \apl{[]} \defi{to braid}
\begin{itemize}
\item \exsen{Bongo bäne bun de mermerae nätɨpnegalle.} \extran{You braided your hair nicely.}
\end{itemize}
\item \anpl{tɨtɨrɨk} \apl{triknen} \defi{to push through, make go through; undo by becoming released, to untie}
\begin{itemize}
\item \exsen{Ngäna torok pop dae toboll de dätäräk.} \extran{I pushed the spear through a hole in the torok plant.}
\item \exsen{Angde gopeyamän walle watta, däräng bo llan a komllaebmae eragwaeya deyatäräkän.} \extran{When he came out of the water, the dog's two ears were unfolded (i.e. they were folded before).}
\end{itemize}
\item \anpl{udaude} \apl{udanen} \defi{to light, start (a fire)}
\begin{itemize}
\item \exsen{Ubi yu de daudeaemeyo.} \extran{Then started some fires.}
\end{itemize}
\item \anpl{ugug} \apl{ugnen} \defi{to make mumu}
\begin{itemize}
\item \exsen{Ngämi ttägäll daugeya.} \extran{We made mumu.}
\end{itemize}
\item \anpl{ullull} \apl{ullnen} \defi{to cross over}
\begin{itemize}
\item \exsen{Ubi tätäm ttongo mo sisor de daulliyu.} \extran{Yesterday, they crossed over a new bridge.}
\end{itemize}
\item \anpl{ume ddäddäl} \apl{ddälnen} \defi{to kiss}
\begin{itemize}
\item \exsen{Bina mokowang lla da gänyan ngäna erem ume bäddäl.} \extran{The man you all want is the one that I will kiss.}
\end{itemize}
\item \anpl{uziz} \apl{[]} \defi{to run}
\item \anpl{wabeb} \apl{wabnen} \defi{to beat, smash, pound}
\begin{itemize}
\item \exsen{Mälla da biye de duwebnegän.} \extran{The woman pounded the taro.}
\item \exsen{Ttängäm mondrog lla da melemang de dällädeyo, gagagäll duwebeyo.} \extran{The gardeners grabbed the servant and beat him badly.}
\end{itemize}
\item \anpl{wallwall} \apl{[]} \defi{yawn}
\item \anpl{wandae} \apl{wandaenen} \defi{to burn}
\item \anpl{wanwen} \apl{[]} \defi{to shake, swing, to shake, swing}
\begin{itemize}
\item \exsen{Bogo bun gowenän ada, "Ddone."} \extran{His head shook like this, "No."}
\item \exsen{Zämllallang a oba bun di dowanaemneyo mikutt me.} \extran{Passers-by shook their heads in anger.}
\end{itemize}
\item \anpl{wanyweny} \apl{[]} \defi{to burn}
\begin{itemize}
\item \exsen{Kullkull wanyweny e gotäbanegnän.} \extran{They were planning to burn a grass fire.}
\end{itemize}
\item \anpl{waswes} \apl{wasnen} \defi{to ask, beg, political group}
\begin{itemize}
\item \exsen{Da ngämi e we bam baweseya, ngäma moko da bongo nangesneg ngämira.} \extran{Whatever we (excl.) ask you for, we want you to do it for us.}
\item \exsen{Ge ngäna ewe de dawes ikopse alle.} \extran{This is what I had asked for in prayer.}
\end{itemize}
\item \anpl{wowo} \apl{wonen} \defi{to clear floating grass by pushing through with a canoe}
\begin{itemize}
\item \exsen{Tätäm ibi tawa de duwonalla.} \extran{Yesterday we cleared the floating grass.}
\end{itemize}
\item \anpl{wɨndwɨnd} \apl{wɨndnen} \defi{to cover, bury}
\begin{itemize}
\item \exsen{Pätt de ekaklle alle dawɨndeya.} \extran{We covered the body with earth.}
\end{itemize}
\item \anpl{yagyeg} \apl{yagnen} \defi{to search, look for}
\begin{itemize}
\item \exsen{Erodias nyongo de deyagnegnän Zon bälle gäz e.} \extran{Herodias was looking for ways to kill John.}
\end{itemize}
\item \anpl{zanggae} \apl{zanggaenen} \defi{to roam, go around}
\begin{itemize}
\item \exsen{Ngäna däba ngata me gonzagae.} \extran{I wandered around in that area.}
\end{itemize}
\item \anpl{zanzi} \apl{[]} \defi{to settle}
\begin{itemize}
\item \exsen{Ge ll} \extran{This boy settled there.}
\end{itemize}
\item \anpl{zaze} \apl{[]} \defi{to give birth, lay, generation}
\begin{itemize}
\item \exsen{Ngäna käp tumang zazeag dan.} \extran{I lay many eggs.}
\item \exsen{ngämo baba bakmall zaze} \extran{my father's siblings (lit. those in the generation with my father)}
\end{itemize}
\item \anpl{zigae} \apl{zigaenen} \defi{to wrap}
\begin{itemize}
\item \exsen{Ngäna mänyän de ttam alle zigae eran.} \extran{I am wrapping the mänyän fish with leaves.}
\item \exsen{Ngämi sana mängalae dazgiaebeya.} \extran{We (excl.) quickly wrapped the sago.}
\end{itemize}
\item \anpl{zizi} \apl{zinen} \defi{to uncover, lift}
\begin{itemize}
\item \exsen{Angde ngämi ttägäll de däziya, duwem gogmam.} \extran{When we (excl.) uncovered the mumu, we ate.}
\item \exsen{Wel gullbe da gongttägän a mägda bo pite de däzinän.} \extran{A strong wind came and lifted his mother's skirt.}
\end{itemize}
\item \anpl{zozo} \apl{zonen} \defi{to rot, go bad}
\begin{itemize}
\item \exsen{Kumddäga ebdo ddɨgatt alle up a mɨnyi bozowän.} \extran{After three days, the banana will rot.}
\item \exsen{Lläkäm a mamall zozoang dan.} \extran{The mushroom goes bad quickly.}
\end{itemize}
\item \anpl{zuwoe} \apl{[]} \defi{to shoot, to pierce; to inject, administer a shot}
\begin{itemize}
\item \exsen{Bodog käbama dallän a sɨmell gullbe de dazuwän.} \extran{Bodog went hunting and shot a boar.}
\item \exsen{Baba bom sana täkäll da nazunan nying me.} \extran{The sago thorn pierced Dad in the leg.}
\item \exsen{Ngänäm doktor da pam alle nazuwan.} \extran{The doctor gave me an injection.}
\end{itemize}
\item \anpl{zämae} \apl{zämaenen} \defi{to pour, put, transfer}
\begin{itemize}
\item \exsen{Ubi ine de zämaenen erallo.} \extran{They are pouring water.}
\item \exsen{Kollba dängälläbeya, nyäng e danzämaeya.} \extran{We bought fish and put them in a bag.}
\end{itemize}
\end{itemize}

\section*{Class II Verbs}
\begin{itemize}
\item \anpl{any} \apl{[]} \defi{to paint}
\begin{itemize}
\item \exsen{Ngäna ngämo mittapa de pällämpälläm palle alle anyan.} \extran{I painted my face with white clay.}
\end{itemize}
\item \anpl{dabän} \apl{dabnen} \defi{to knock a leaf, e.g. a tobacco leaf}
\item \anpl{dagal} \apl{dadeg} \defi{to board, get on, to board, load, put on}
\begin{itemize}
\item \exsen{Ngämi deyareya gall e godagaleya.} \extran{We went to the boat and got on.}
\item \exsen{Ngäna gongäs gall e godagal.} \extran{I returned to the boat and got on.}
\item \exsen{Ngämo män kälsre de gall e dädagal.} \extran{I put my daughter onto the boat.}
\end{itemize}
\item \anpl{ddaebän} \apl{ddaebnen} \defi{to divorce, to adopt}
\begin{itemize}
\item \exsen{Ngäna mɨnyi mälla de bäddebän.} \extran{I will divorce my wife.}
\item \exsen{Ngäna Jepet bom ngämlle llɨg de däddebän.} \extran{I adopted Jepet as my son.}
\end{itemize}
\item \anpl{ddogoll} \apl{ddäganen} \defi{to put together, to stick, to join, to lean}
\begin{itemize}
\item \exsen{Angde tine pittnen a gottamänän, dibaballe ddäganen de dängkam.} \extran{After the sago leaf weaving finished, I started putting together the roof.}
\item \exsen{Lla da päre abo gobällan kottllam bo gollob pallkepallke de dänglläbeyo a kottllam bo patme däddäganeyo.} \extran{Then the people sadly went to the turtle, collected all the pieces of his shell, and put him back together.}
\item \exsen{Tongle da addgollan.} \extran{The leech is stuck.}
\item \exsen{Ngäna goddägol.} \extran{I joined.}
\item \exsen{Llo pätt me ddogollag dan.} \extran{It's leaning on the tree trunk.}
\end{itemize}
\item \anpl{ddällbän} \apl{ddäddälläb, ddälläbnen} \defi{uproot}
\item \anpl{ddän} \apl{[]} \defi{to pick, gather, harvest}
\begin{itemize}
\item \exsen{Dädme ddob wayati de däddaebeya.} \extran{We harvested some watermelons there.}
\end{itemize}
\item \anpl{däbäll} \apl{[]} \defi{to touch}
\begin{itemize}
\item \exsen{Nädbällnegan.} \extran{He touched him.}
\item \exsen{Nädbällnegan} \extran{He touched three people at different times.}
\item \exsen{Lla da ubim yadbällan.} \extran{He touched three people at the same time.}
\end{itemize}
\item \anpl{dämen} \apl{dämadäme, dämanen} \defi{to sit}
\begin{itemize}
\item \exsen{Ngäna dämenang dan.} \extran{I am sitting.}
\item \exsen{Adämeneyo!} \extran{(You two) sit down!}
\item \exsen{Ttongo lla da dämenma toko me adämenan.} \extran{A person sat on top of the chair.}
\end{itemize}
\item \anpl{eräl} \apl{[]} \defi{to name}
\item \anpl{gabän} \apl{gabnen} \defi{to cut in half or to split, e.g. to split a deer thigh with someone.}
\item \anpl{gambän} \apl{[]} \defi{to stand}
\item \anpl{gany} \apl{ganen} \defi{to plant, place in the ground, put in, to focus, tune}
\begin{itemize}
\item \exsen{Angde pos ganen a gottamänän ngäna sipel gog.} \extran{When the post planting finished, I took a rest.}
\item \exsen{Llan de ngämi däganyeya eka we.} \extran{We tuned our ears for the sound.}
\end{itemize}
\item \anpl{gazen} \apl{gez, gazgez, gaznen} \defi{to come out, get out, exit, escape, to rise, come out; shine, to take out}
\begin{itemize}
\item \exsen{Nag wiya agezän.} \extran{Friend, come out.}
\item \exsen{Bogo ma ik atta gogezänän wälläng e dindugän.} \extran{He got out from inside and ran into the bush.}
\item \exsen{Kandärmang, yu da daollemae llamda da bälle gazenma da ddone mullae gogon.} \extran{Sadly, the fire got closer to the old man and he didn't have a way he could escape.}
\item \exsen{Yäbäd a mermer abal gogezänän.} \extran{The sun came out, shining nicely.}
\item \exsen{Kok a gazen allan.} \extran{The moon is rising.}
\item \exsen{Käsre tos de dägazen nyäng ik att de a batri käp de dazer.} \extran{Then I took the torch out from the bag and put in the batteries.}
\item \exsen{Aeya obom bigezänän?} \extran{Who will come take him out?}
\end{itemize}
\item \anpl{gäbän} \apl{gägäb} \defi{to jump, hop}
\begin{itemize}
\item \exsen{Gäbänmällang dan.} \extran{He is hopping.}
\item \exsen{Män a gall att agäbänan.} \extran{The girl jumped out of the canoe.}
\end{itemize}
\item \anpl{gädän} \apl{gädaenen} \defi{to land}
\item \anpl{gälbän} \apl{gälbänan, gägleb, gälbnan} \defi{to knock (a fruit), to defeather, depilate, remove hair; remove teeth}
\begin{itemize}
\item \exsen{Polle ma ik i angde gozenän, up wo gälbänanatt de ikop dägagän.} \extran{When he went inside the fence, he saw the fallen ripe banana.}
\end{itemize}
\item \anpl{gällbän} \apl{[]} \defi{to break}
\begin{itemize}
\item \exsen{Ngämo llimba da agällbänan.} \extran{My fingernail broke.}
\end{itemize}
\item \anpl{ibeny} \apl{ibeb, ibenen} \defi{to plant}
\begin{itemize}
\item \exsen{Bogo ibeny eran.} \extran{He is planting.}
\item \exsen{Biye de bongo ai dan ibik alle nibe.} \extran{You can plant taro with the ibik stick.}
\item \exsen{Ttängäm e gongttägeyo, misdae up daebegaeya ibebatta.} \extran{They arrived at the garden; only bananas were planted there.}
\end{itemize}
\item \anpl{idän} \apl{idäd, idnen} \defi{to pick, harvest; dig up, uproot}
\begin{itemize}
\item \exsen{Ubi mɨnyi mätta de bidaebneyo.} \extran{They will dig up the yams and bring them over.}
\item \exsen{Ada wayati da moko alle dädeyo.} \extran{[You] pick watermelons as you please.}
\end{itemize}
\item \anpl{inggol} \apl{inggoemeny, inggolnen, inggoenen, inggolmäll, inggoemäll} \defi{to move}
\begin{itemize}
\item \exsen{Mälla da angde ikop dägagän, llɨg bo dowae e guinggolän.} \extran{When the woman saw it, she moved closer to the boy.}
\end{itemize}
\item \anpl{irängän} \apl{irnen} \defi{to come out, to get out, lift out}
\begin{itemize}
\item \exsen{Lla da ine ik atta guirängänän.} \extran{The man come out from the water.}
\item \exsen{Angde ngäna net de tuk i dirängän, käza da net ik i gozenän.} \extran{When I lifted the net into the air, a crocodile went inside it.}
\item \exsen{Mana ddone mullae gogon obo känyer, ede adawatta bogo Lama bälle wälle gogon, Emi bälle yirngän e.} \extran{Mana wasn't able to herself, so she called for Lama to lift Emi out.}
\end{itemize}
\item \anpl{ittall} \apl{ittaenen} \defi{to hang, to follow closely, stalk (+ peyang)}
\begin{itemize}
\item \exsen{Kaptte da ittananang da pätanan e.} \extran{The clothes are hanging to dry.}
\item \exsen{Mälla da net de mɨnyi bonyän baittallän.} \extran{The woman will take the net and hang it.}
\item \exsen{Molemoleg a llo ttam me itraenen anggan.} \extran{Caterpillars hang from tree leaves.}
\item \exsen{Ngäna Wagibo bo peyang buittall.} \extran{I will stalk Wagibo.}
\end{itemize}
\item \anpl{kakal} \apl{kakek} \defi{to enter, board, go in, load}
\begin{itemize}
\item \exsen{Bogo gall e gokakalän.} \extran{She got in the canoe.}
\item \exsen{Ngäna sɨmell lelang atta llo we okakalan.} \extran{I was afraid of the pig, so I went in the tree.}
\end{itemize}
\item \anpl{koen} \apl{koemeny} \defi{to turn back}
\begin{itemize}
\item \exsen{Ngäna llowam atta koen allan.} \extran{I'm turning back due to fatigue.}
\end{itemize}
\item \anpl{kolldän} \apl{kolldonen} \defi{to shoot; stab}
\begin{itemize}
\item \exsen{Sɨmell de enanae dukollädnegän pakos alle.} \extran{He shot the pig straight through with the arrow.}
\end{itemize}
\item \anpl{kollwany} \apl{kollwanen} \defi{to hang}
\begin{itemize}
\item \exsen{Pakätt de llo ngoeang me dukullwenyallo.} \extran{They hung the robe on the tree branch.}
\end{itemize}
\item \anpl{kän} \apl{känan} \defi{to withdraw, come out, to remove, take out, take off, undo}
\begin{itemize}
\item \exsen{Ngäna skul atta gongkän greid paeb me.} \extran{I withdrew from school in grade five.}
\item \exsen{Bäne sod de nängkän.} \extran{Take off your shirt.}
\item \exsen{Ngäna net de apte tär nängkänan.} \extran{I undid the string on one side of the net.}
\item \exsen{Gänyme nge kopnen ma patt de aeya dängkänän?} \extran{Who took out the coconut dehusker (from the ground) here?}
\end{itemize}
\item \anpl{lläpän} \apl{lläplläp, lläpnan} \defi{to dig, harvest, unearth (a tuber or corm)}
\begin{itemize}
\item \exsen{Mälla da dagaeya ttängäm e abällan mätta lläplläp e.} \extran{The women went to the garden to dig the yams.}
\item \exsen{Bogo biye kemibi dälläpnegalle.} \extran{He used to harvest many taro roots.}
\end{itemize}
\item \anpl{llɨtɨn} \apl{llɨtmäll} \defi{to select}
\item \anpl{läpon} \apl{läponen} \defi{to be amazed, be in awe, to doubt}
\begin{itemize}
\item \exsen{Ubira tämamae lla yaba läponang gogon.} \extran{All of them were amazed.}
\item \exsen{Ubi daläponeyo.} \extran{They were amazed.}
\end{itemize}
\item \anpl{mamon} \apl{manen} \defi{to string (e.g. a bow, sago beater), to fashion, shape, make}
\begin{itemize}
\item \exsen{Tätäm ngäna bägäl de damon.} \extran{Yesterday, I strung the bow.}
\item \exsen{Bägäl a zime amonan?} \extran{Is the bow strung?}
\item \exsen{Ttäle da amonan.} \extran{My leg is strung (i.e. it's tight and I can't flex it).}
\item \exsen{Mälla da mɨnyi abor de bamonän.} \extran{The woman will fashion a sago beater.}
\end{itemize}
\item \anpl{ngällbän} \apl{ngällbänen, ngängälläb, ngällbnan} \defi{to rise, arise, come up, to awake, wake up, get up, to lift, to get, take, to wake up, to wake}
\begin{itemize}
\item \exsen{Ine ulle da gongälllbänmällnän.} \extran{The big wave was rising.}
\item \exsen{Ddobag a gudae ngällbaenen eran yunu atta.} \extran{Some wake up from sleep early.}
\item \exsen{Kottllam de kapu wi dängällbäneyo.} \extran{They lifted the turtle up to carry.}
\item \exsen{teks mani ngällbänang lla} \extran{tax collector}
\item \exsen{Dindugän, do täl pallkoll dingällbänän.} \extran{He ran and got a piece of bamboo.}
\item \exsen{Ngäna bam danglläbän.} \extran{I woke you up.}
\end{itemize}
\item \anpl{nyäny} \apl{nyänan} \defi{to paint, to anoint}
\begin{itemize}
\item \exsen{Ubi pale alle nänyallo.} \extran{They painted themselves with white clay.}
\item \exsen{Ubi mer mollong owel de dällädaebeyo obo pätt nyäny e.} \extran{They bought fragrant oils for anointing his body.}
\end{itemize}
\item \anpl{otal} \apl{otaenen} \defi{to perch}
\begin{itemize}
\item \exsen{Pa da llo me gotalalle.} \extran{The bird perched in the tree.}
\end{itemize}
\item \anpl{pentngeny} \apl{[]} \defi{to trip}
\begin{itemize}
\item \exsen{Bogo säremsärem gompetngenyän yu pallkoll me.} \extran{In the darkness, he tripped over a piece of wood.}
\end{itemize}
\item \anpl{pirngän} \apl{pirpir, pirnen} \defi{to draw a weapon, take out}
\begin{itemize}
\item \exsen{Ngäna pakos de dapirngän.} \extran{I took out a spear.}
\end{itemize}
\item \anpl{pllaengän} \apl{[]} \defi{strike}
\item \anpl{pällttän} \apl{päpällätt, pllättnan} \defi{to set off, start walking}
\begin{itemize}
\item \exsen{Däpällätt abo do Taolang.} \extran{I started walking to Taolang.}
\end{itemize}
\item \anpl{pängän} \apl{[]} \defi{to disappear}
\begin{itemize}
\item \exsen{Bogo apngänän.} \extran{He disappeared.}
\end{itemize}
\item \anpl{pätangeny} \apl{[]} \defi{splash}
\item \anpl{pätaräll} \apl{[]} \defi{splash}
\item \anpl{tergony} \apl{[]} \defi{to spread, unfold}
\begin{itemize}
\item \exsen{Lla da tater de mɨnyi bitergonyeyo.} \extran{The people will spread mats.}
\end{itemize}
\item \anpl{togol} \apl{tägatäge, täganen} \defi{to hide; go away to have sex secretly, to hide}
\begin{itemize}
\item \exsen{Ngäna ngattong botägol.} \extran{I will hide first.}
\item \exsen{Bongo zime atägolalle?} \extran{Are you already hiding?}
\item \exsen{Ngämim deyatogolmällneyo ngäma mäg.} \extran{Our (excl.) mothers were hiding us.}
\end{itemize}
\item \anpl{tomon} \apl{tomonen} \defi{to wait, to wait for, await}
\begin{itemize}
\item \exsen{Ngäna kälakälae gontämon do yäbäd a angde goklowän dam ngäna ibi de gongkam.} \extran{I waited a little until the sun set; then I started walking.}
\item \exsen{Ngäna mänmän e bantmonneg.} \extran{I'll wait for the girls.}
\item \exsen{Bogo obom deyantämonän.} \extran{She awaited him.}
\end{itemize}
\item \anpl{ttaengän} \apl{ttaengnen} \defi{to pull (a plant sucker)}
\item \anpl{ttäpen} \apl{[]} \defi{to snap, break, to break, tear (esp. something long), split}
\begin{itemize}
\item \exsen{Kwalde mäse pakos de gonttalägän sämell zuwoe e, be bägäl wadär a gottäpenän.} \extran{Kwalde tried to aim the arrow to shoot the pig, but the bow string snapped.}
\item \exsen{Ngäna ngämo bunkom de nättpenan.} \extran{I broke a strand of my hair.}
\item \exsen{Obo tudi di kollba da dättpenän.} \extran{The fish broke her fishing line.}
\end{itemize}
\item \anpl{tärpan} \apl{[]} \defi{cut}
\item \anpl{täräpän} \apl{tatäräp, täräpnan} \defi{to cut}
\item \anpl{wamän} \apl{[]} \defi{to go out, dissipate, extinguish}
\begin{itemize}
\item \exsen{Yu da dowamänän.} \extran{The fire went out.}
\end{itemize}
\item \anpl{yattän} \apl{[]} \defi{to disembark, get off, get out}
\begin{itemize}
\item \exsen{Gall atta goyattänän.} \extran{She got out of the canoe.}
\end{itemize}
\item \anpl{yatän} \apl{[]} \defi{to fetch (water)}
\begin{itemize}
\item \exsen{Oblle ine neyatän.} \extran{[You] fetch water for him.}
\end{itemize}
\item \anpl{yäbäd gazen} \apl{[]} \defi{to dawn}
\end{itemize}

\section*{Class III Verbs}
\begin{itemize}
\item \anpl{ballängg} \apl{[]} \defi{to predict, foresee, prophesize, to greet, welcome}
\item \anpl{ballɨngg} \apl{ballɨminy} \defi{to welcome, greet, to predict}
\begin{itemize}
\item \exsen{Ngänaeka alle yamballägallo.} \extran{In tears, they greeted him.}
\item \exsen{Sisor pazi di bamballmenyneyo.} \extran{They will be celebrating the new year.}
\item \exsen{Bogo abade ngasnen e ttoen de damballmenynegän.} \extran{He predicted things that will happen in the future.}
\end{itemize}
\item \anpl{beyambäg} \apl{beyambmeny} \defi{to chase}
\begin{itemize}
\item \exsen{Ngäna sɨmell beyambägag dan.} \extran{I am a pig-chaser.}
\item \exsen{Yäbäd imneimne gumbiebägän ddone dingmenän.} \extran{Sun pursued from behind, but did not reach him.}
\end{itemize}
\item \anpl{binzeg} \apl{binzenen} \defi{to heat, warm}
\item \anpl{bombllag} \apl{[]} \defi{to find}
\begin{itemize}
\item \exsen{Ngäna mɨnyi obom bombllag.} \extran{I will find him.}
\item \exsen{Bogo mɨnyi obom bombllagän.} \extran{He will find him.}
\end{itemize}
\item \anpl{bungg} \apl{[]} \defi{to ambush}
\item \anpl{bällma ontog} \apl{ontomeny} \defi{to spit at}
\begin{itemize}
\item \exsen{Ubi obom bällma dawäntmenyneyo.} \extran{They spat at him.}
\end{itemize}
\item \anpl{bällängg} \apl{bällämeny} \defi{to split up, separate}
\begin{itemize}
\item \exsen{Ubi gombällägmenyamän.} \extran{They were splitting up.}
\end{itemize}
\item \anpl{bändaeg} \apl{bändaemeny} \defi{to pull an all-nighter (stay awake)}
\begin{itemize}
\item \exsen{Ibi iddob mɨnyi beyambadägeya.} \extran{We (incl.) will pull an all-nighter.}
\end{itemize}
\item \anpl{dangg} \apl{dameny} \defi{to burn}
\begin{itemize}
\item \exsen{Nyongo da andegan.} \extran{The road is burning.}
\end{itemize}
\item \anpl{ddollombog} \apl{[]} \defi{to misspeak, speak with mistakes}
\begin{itemize}
\item \exsen{Tame eka de da ngäna bäpany, da erame ngäna banddollbog, ede mäzi ngänäm ttättle bageyo.} \extran{When I speak Taeme, whenever I make mistakes, they will correct me.}
\end{itemize}
\item \anpl{ddungg} \apl{dduminy} \defi{to decapitate}
\begin{itemize}
\item \exsen{Zon bom inkätt danddugän.} \extran{He decapitated John.}
\end{itemize}
\item \anpl{ddällombog} \apl{ddällombomeny} \defi{to miss}
\begin{itemize}
\item \exsen{Bogo nyongo de nanddällbogan.} \extran{He missed the road.}
\item \exsen{Bogo angde tuk i nallnan mo de nanddällbogan ttongo.} \extran{She missed a step going up the ladder.}
\end{itemize}
\item \anpl{dumbeg} \apl{[]} \defi{give everything}
\item \anpl{dämbaemeny} \apl{[]} \defi{to doze off}
\begin{itemize}
\item \exsen{Dagirnän e: yunu gondäbaemenynän.} \extran{and stayed there for a long, long time until he dozed off.}
\end{itemize}
\item \anpl{dämenangg} \apl{dämameny} \defi{causative-applicative form of dämen, applicative form of dämen}
\begin{itemize}
\item \exsen{Nändämenanggan.} \extran{He sat him down.}
\item \exsen{Ngäna Kate bom täräb ma me dändämenang.} \extran{I sat down with Kate at the funeral.}
\end{itemize}
\item \anpl{dämoengg} \apl{[]} \defi{causative-applicative form of dämoe}
\begin{itemize}
\item \exsen{Ngäna bam sɨmell ikop e dandämoengg.} \extran{I sent you to see the pig.}
\end{itemize}
\item \anpl{därängg} \apl{[]} \defi{to lead}
\begin{itemize}
\item \exsen{Mälla da obom ttängäm att de ngattong dändrägän menae e.} \extran{The woman led him outside the village.}
\end{itemize}
\item \anpl{eka tameny} \apl{[]} \defi{to discuss, converse, to tell; converse with, speak to}
\begin{itemize}
\item \exsen{Ngämo nagnag aba kämall eka gontemenyne.} \extran{I was conversing with my friends.}
\item \exsen{Ubi ngämim gänya eka walle mae eka deyantemenynegnän.} \extran{They were only speaking to us in this language.}
\end{itemize}
\item \anpl{eka ttängkɨg} \apl{ttängkɨminy} \defi{to answer someone with force, to talk against someone}
\item \anpl{erongg} \apl{eromeny} \defi{to start weaving (crossing strings)}
\begin{itemize}
\item \exsen{Essie nga wiya-o, bongo ngämo naerongg tater de.} \extran{Essie, you come and start weaving my mat.}
\end{itemize}
\item \anpl{erängg} \apl{ermeny} \defi{to test, try; taste}
\begin{itemize}
\item \exsen{Tos de dägazen nyäng ik att de a batri käp de dazer. Angde daerängg mermer gogllayän.} \extran{I took out the torch from inside my bag and put in the batteries. When I tested it, it properly shone.}
\item \exsen{Be ami ddäddägeyo, mokwang abal daeränggeyo.} \extran{But those who tried eating it found that it tasted delicious.}
\end{itemize}
\item \anpl{gambenängg} \apl{[]} \defi{to help (someone) stand}
\item \anpl{gambäg} \apl{gambämeny} \defi{to stand closely}
\item \anpl{gameny} \apl{[]} \defi{to become cloudy, when dark rain clouds are forming}
\item \anpl{gollaembäg} \apl{gollaembmeny} \defi{applicative form of gollab}
\begin{itemize}
\item \exsen{Mänmän mängalae ine walle danggollaebägeyo.} \extran{The girls quickly poured water on someone.}
\end{itemize}
\item \anpl{gonangg} \apl{gonameny} \defi{to cook}
\item \anpl{gongg} \apl{gomeny} \defi{to disturb a bees nest and then to feel the bites}
\item \anpl{gungg} \apl{guminy} \defi{to marry a widow}
\begin{itemize}
\item \exsen{Angde Kwalde kuddäll agan, Kidarga ako mosen da bäne mik di nanggugan.} \extran{When Kwalde died, Kidarga married his older brother's (i.e. Kwalde's) widow.}
\end{itemize}
\item \anpl{gämbanengg} \apl{[]} \defi{causative-applicative form of gambän}
\begin{itemize}
\item \exsen{Ngäna bam danggebänängg.} \extran{I made you stand.}
\end{itemize}
\item \anpl{gämäll tänggag} \apl{[]} \defi{to steal}
\begin{itemize}
\item \exsen{Da bongo ddone otät de nonttog, bam mɨnyi gämäll nantägegän.} \extran{If you do not give him food, he will steal.}
\end{itemize}
\item \anpl{gɨngg} \apl{gɨminy} \defi{to ambush, gang up on, attack in a large group}
\begin{itemize}
\item \exsen{Nogat bom ubony a ere daeränggeyo, gudae enanae dɨnggɨgeyo.} \extran{When bees started biting Nogat, they really attacked him.}
\end{itemize}
\item \anpl{indugoeg} \apl{[]} \defi{to command}
\begin{itemize}
\item \exsen{Bongo gagäll anyke, ngäna bam indugoeg nallan, agezän obo guwo watta.} \extran{You evil spirit, I command you to come out of his heart.}
\item \exsen{Mägda män de naindugoegan kaptte mättmätt e.} \extran{The girl's mother commanded her to put on clothes.}
\end{itemize}
\item \anpl{indurgoeg} \apl{indurgoemeny} \defi{to yell}
\item \anpl{inggoemeny} \apl{[]} \defi{to joke, to blaspheme; insult, mock}
\begin{itemize}
\item \exsen{Ende walle inggoemeny amalla.} \extran{We are making jokes in Ende.}
\item \exsen{Ngäna Ingglis eka walle de eka bogne, buinggoemenyne.} \extran{I will speak in English, making jokes.}
\item \exsen{Bogo Adi bom inggoemeny eran.} \extran{He is blaspheming God.}
\item \exsen{Sabi tamenyang a inggoemeny eka de däpanyaemneyo.} \extran{The teachers of the law were saying insults.}
\end{itemize}
\item \anpl{inttemongg} \apl{inttemeny} \defi{to let go}
\item \anpl{inttemängg} \apl{inttoemeny} \defi{to part ways, leave, to leave, see off, release, set free}
\begin{itemize}
\item \exsen{Eka tamenyatt me, ubi guinttemänggeyo a oba ma we deyareyo.} \extran{After discussing, they parted ways and went to their own homes.}
\item \exsen{Kottllam bom dätrameyo bem ulle we dainttemonggeyo.} \extran{They carried the turtle to the ocean and set him free.}
\end{itemize}
\item \anpl{inttoemängg} \apl{inttoemmeny} \defi{to see (someone) off}
\item \anpl{kallkäll gllaengg} \apl{gllaemeny} \defi{to warm up, get warm, warm oneself}
\begin{itemize}
\item \exsen{Bogo yu me kallkäll gonggllaemenynän.} \extran{He was warming up by the fire.}
\item \exsen{Da ngäna ddone yu bipellknegalle bablle, bongo ddone gänya mer yu mi kallkäll bonggllaegalle.} \extran{If I hadn't chopped wood for you, you couldn't get warm by this nice fire.}
\end{itemize}
\item \anpl{kalläntäg} \apl{kalläntämeny} \defi{to split}
\begin{itemize}
\item \exsen{Bogo mätta ulleulle de deyangkälltägalle.} \extran{She split the big yams.}
\end{itemize}
\item \anpl{kallɨngg} \apl{[]} \defi{to kill}
\begin{itemize}
\item \exsen{Sɨmell de mɨnyi bangkallɨgiyu.} \extran{They will kill the pig.}
\end{itemize}
\item \anpl{kambäg} \apl{kambämäll} \defi{to escape, flee}
\begin{itemize}
\item \exsen{Ngäna gänyeri angkebägan.} \extran{I fled this way.}
\end{itemize}
\item \anpl{kangkäg} \apl{[]} \defi{to carry, bear (a load)}
\item \anpl{komengg} \apl{[]} \defi{to go and come back}
\item \anpl{komongg} \apl{[]} \defi{causative-applicative form of ony (to make carry; bring for, take to; take from)}
\begin{itemize}
\item \exsen{Grace ngänäm gull de dangkomonggän.} \extran{Grace had me bring the net.}
\item \exsen{Bong obo pope sɨmell midd de dangkomonggän.} \extran{Bong brought the pork for his uncle.}
\item \exsen{Ttall a obo za de dangkämonggän.} \extran{The wallaby took her thing.}
\end{itemize}
\item \anpl{kälämpag} \apl{kälämpameny} \defi{to pad (with cloth or leaves to support a heavy load), to patch (e.g. roof)}
\item \anpl{kämbäg} \apl{kämbämeny, kämbmeny} \defi{to dive, to baptize}
\begin{itemize}
\item \exsen{Ngämi ine kämbmeny de dängkameya.} \extran{We (excl.) started diving.}
\item \exsen{Wiya angkäbäg!} \extran{[You] come dive!}
\item \exsen{Zon ine kämbägag} \extran{John the Baptist}
\item \exsen{Pasta da obom ine dängkäbägän dedme.} \extran{The pastor baptized him there.}
\end{itemize}
\item \anpl{kämongg} \apl{kämomeny} \defi{to get from someone}
\item \anpl{lläntäg} \apl{lläntmeny} \defi{to tell}
\begin{itemize}
\item \exsen{Golläntmenyän tämamae ngalen de.} \extran{He told him everything.}
\end{itemize}
\item \anpl{menttlog} \apl{menttlomeny} \defi{to agree}
\item \anpl{menttäg} \apl{menttämeny} \defi{to shoulder, carry one one's shoulders or head}
\begin{itemize}
\item \exsen{Nyäng goimänttmenyeyo deyarneyo abo ma we.} \extran{They put the baskets on their shoulders and headed home.}
\end{itemize}
\item \anpl{molle dungg} \apl{dumeny} \defi{to smell, take a whiff}
\item \anpl{mälanzeg} \apl{mälanzemeny} \defi{to tie}
\item \anpl{ngallmeny} \apl{[]} \defi{to advise}
\item \anpl{nganzig} \apl{nganziminy} \defi{to pass, overtake}
\begin{itemize}
\item \exsen{Ddia da angde inu kuddäll gogon, kottllam a deyanzigän.} \extran{While Deer was dead sleep, Tortoise passed right by him.}
\end{itemize}
\item \anpl{ngarängg} \apl{ngarmäll} \defi{to encounter, meet, run into}
\begin{itemize}
\item \exsen{Da ngäna lla de bangerängg nyongo me, mɨnyi Ende eka walle eka bäntameny.} \extran{If I encounter someone on the street, I will speak in Ende to them.}
\item \exsen{Gänya ngata me ddädddäg de ddone dangerängg.} \extran{I didn't encounter any animals at that spot.}
\end{itemize}
\item \anpl{ngllongg} \apl{ngllomeny} \defi{to forget}
\begin{itemize}
\item \exsen{Ubi gongllomenynegän otät kongkom e.} \extran{They forgot to bring food.}
\end{itemize}
\item \anpl{ngonongg} \apl{ngonomeny} \defi{to think, to think about, think of, recall, remember}
\begin{itemize}
\item \exsen{Bogo ada gongnomenynän ada ka däräng a pollnenang me erallo.} \extran{He was thinking that the dogs were barking.}
\item \exsen{Emi ulle abal gongnomenyän.} \extran{Emi was thinking hard.}
\item \exsen{Ngäna bam angnonggan.} \extran{I thought of you.}
\item \exsen{Eka de tämamae ngäna era ddone ngonomeny allan.} \extran{I can't recall all the words.}
\end{itemize}
\item \anpl{ngämingg} \apl{ngämiminy} \defi{to help, to answer}
\begin{itemize}
\item \exsen{Lla da ttongo lla de nangminggan towall ttokoe e.} \extran{A man helps another man to cut grass.}
\item \exsen{Ngäna bam mɨnyi bangmingg.} \extran{I will help you.}
\item \exsen{Ibi mɨnyi iba mälla de bangmiminyaemnalla.} \extran{We (incl.) will be supporting our women.}
\item \exsen{Bogo do gungminggän ada, "Aoǃ"} \extran{There he answered, "Yesǃ"}
\end{itemize}
\item \anpl{ngänaeka pollongg} \apl{pollomeny} \defi{to burst into tears}
\begin{itemize}
\item \exsen{Bogo ngänaeka gompllogon.} \extran{He burst into tears.}
\end{itemize}
\item \anpl{ngänongg} \apl{ngänomeny} \defi{to think about, think of, believe}
\begin{itemize}
\item \exsen{Ngäna bam angnonggan.} \extran{I thought of you.}
\end{itemize}
\item \anpl{ngänttäg} \apl{llɨtaem} \defi{to arrive, return, to bring, carry}
\begin{itemize}
\item \exsen{Ge lla ngänttäg allan ttongo ttängäm atta kilikiliang.} \extran{This man is returning from another place happily.}
\item \exsen{Ngämo näkäp e gongttägän.} \extran{It came to my mind.}
\item \exsen{Mɨnyi ddob sisor eka da ako bälltaemnän.} \extran{Then other new languages will arrive.}
\item \exsen{Llɨg kälsre de ine da do ddob llayaba pate dängttägän.} \extran{The water carried the boy over to some people over there.}
\end{itemize}
\item \anpl{paengg} \apl{paemeny} \defi{to guess}
\begin{itemize}
\item \exsen{Ngäna paengg eran ada Jugu mɨnyi sisri bongttägän Wim atta.} \extran{I'm guessing that Jugu will arrive from Wim now.}
\end{itemize}
\item \anpl{pallängkmeny} \apl{[]} \defi{to divide, to judge}
\begin{itemize}
\item \exsen{Da ttongo kantri me lla da bompallängkmenynegän obaoba komlla kullum e a bogäddeyo obaoba, däbe kantri da mɨnyi ddone bakällmewän.} \extran{If the people in a country divide themselves into two groups and fight each other, that country will not survive.}
\item \exsen{God bo enanae ttoen pallängkmeny} \extran{God's Final Judgment}
\end{itemize}
\item \anpl{pameny} \apl{[]} \defi{to scream}
\begin{itemize}
\item \exsen{Lla da ompamenyan.} \extran{The man screamed.}
\end{itemize}
\item \anpl{penangg} \apl{penameny} \defi{to set on fire}
\begin{itemize}
\item \exsen{Lla da llo de penangg erallo.} \extran{People set fire to the trees.}
\end{itemize}
\item \anpl{pendäg} \apl{pendämeny} \defi{to trip, to push to the ground, jostle, trip}
\begin{itemize}
\item \exsen{Gopendägalle ekaklle we.} \extran{He was tripping and falling to the ground.}
\item \exsen{Lla gul ulle daeya, Yesu adame obo kollmällang de umllang dägnegän oblle gall kälsäre särämbae e, oba lla da ddone obom beyawändägallo a bimpidägmällnallo.} \extran{The crowd was big, so Jesus told his disciples to prepare a small boat for him so that people would not come crowd him and be pushing him.}
\end{itemize}
\item \anpl{pengg} \apl{[]} \defi{to make the killing shot, deliver the final blow}
\begin{itemize}
\item \exsen{Ddäg aeya nampegan obo ddäg dan.} \extran{The back [of an animal] belongs to him, the one who delivered the final blow.}
\end{itemize}
\item \anpl{penongg} \apl{penomeny} \defi{to burn, set on fire, ignite, to burn, set on fire}
\begin{itemize}
\item \exsen{Ge towall a mängalae penongg allan.} \extran{This grass is burning fast.}
\item \exsen{Yäbäd a aempononggan.} \extran{The sun blazed.}
\item \exsen{Ngäna towall penameny gongkam.} \extran{I started to ignite the grass.}
\item \exsen{Yu daempononggeyo ma de.} \extran{They set fire to the house.}
\item \exsen{Ause da to dänglläbänän, kullkull de daimponomenyän tämamae.} \extran{The old woman got the light and burned the whole garden.}
\end{itemize}
\item \anpl{peraingg} \apl{peraiminy} \defi{to cross, intersect, to protrude, stick out}
\begin{itemize}
\item \exsen{Tawe da goimpregeyo.} \extran{The tawe palms were crossed.}
\item \exsen{Ngäna Karama atta yaran, ngämi Winson alle aimpregalla.} \extran{I came back from Karama; Winson and I had intersecting paths (but we didn't see each other).}
\item \exsen{Sɨmell ada daeya, ngoi peraingg att.} \extran{The pig, its tusks were sticking out like this.}
\end{itemize}
\item \anpl{pinsäg} \apl{[]} \defi{to tear}
\begin{itemize}
\item \exsen{Iddpo da komlla gumpisägeyo tuk alle do matu.} \extran{The cloth tore in two from top to bottom.}
\end{itemize}
\item \anpl{pitraempäg} \apl{[]} \defi{to make a failed shot, shoot an arrow that falls}
\item \anpl{plengg} \apl{plämeny} \defi{to die, to cut down, cut off}
\begin{itemize}
\item \exsen{Bogo däplewän.} \extran{He died.}
\item \exsen{Ngäna llo de plengg eran.} \extran{I am cutting down the tree.}
\end{itemize}
\item \anpl{poloengg} \apl{poloemeny} \defi{to insult}
\item \anpl{poper ingkoeg} \apl{ingkoemeny} \defi{to scare, startle, frighten}
\begin{itemize}
\item \exsen{Poper naengkoegan.} \extran{He scared her.}
\item \exsen{Mänmän de poper inkoegag bongttägän.} \extran{He will arrive suddenly, startling the girls.}
\end{itemize}
\item \anpl{poper kwingg} \apl{[]} \defi{to surprise}
\begin{itemize}
\item \exsen{Ngäna obom poper nangkwigan.} \extran{I surprised her.}
\end{itemize}
\item \anpl{pungg} \apl{puminy} \defi{to get sick, be in pain, to take the first chop with the axe}
\begin{itemize}
\item \exsen{Da bam kungge da naddägän, bongo mɨnyi ttattlle ampug.} \extran{If a spider bites you, you will get very sick.}
\end{itemize}
\item \anpl{pälengg} \apl{[]} \defi{to drop down, fall}
\begin{itemize}
\item \exsen{Angde gogllaeaebne, yogoll ulle da dipliwän.} \extran{While we were paddling, a heavy rain started coming down.}
\end{itemize}
\item \anpl{pänaemeny} \apl{[]} \defi{to check}
\begin{itemize}
\item \exsen{Mälla da mɨnyi bobällnän mätta mɨnyi bampnaemenyaemneyo ada mätta da zäme käp agan abo.} \extran{The women will go and check the yams, whether they are ready yet.}
\end{itemize}
\item \anpl{pänaengg} \apl{[]} \defi{to move opposite}
\item \anpl{pänddäg} \apl{pänddmeny, pänddägmeny} \defi{to start a grassfire, to hatch, to break water, to mature (of a plant or a person)}
\begin{itemize}
\item \exsen{Bogo gompäddägän.} \extran{He started the grassfire.}
\item \exsen{Pauro käp a gompäddägän.} \extran{The chicken egg started hatching.}
\end{itemize}
\item \anpl{pänddäg} \apl{pänddmeny} \defi{causative-applicative form of päddpädd}
\begin{itemize}
\item \exsen{Lla da täl de dämpäddmenyaemallo yu alle.} \extran{People used to blow up bamboo with fire.}
\end{itemize}
\item \anpl{pänggmeny} \apl{[]} \defi{to protect, look after, take care of}
\begin{itemize}
\item \exsen{Ede adawatta ibi mɨnyi iba pätt de mermerae bampägmenyaemnalla.} \extran{This is why we must take care of our bodies.}
\end{itemize}
\item \anpl{pängmeny} \apl{[]} \defi{to scope out}
\item \anpl{pänongg} \apl{[]} \defi{to awake, wake up, get up, to wake up, wake}
\begin{itemize}
\item \exsen{Ngäna obom pänongg eran.} \extran{I am waking him up.}
\end{itemize}
\item \anpl{pänyanz} \apl{pänyanznen} \defi{to grow}
\begin{itemize}
\item \exsen{Saly ddone ada lla tupi da pänyanz eran.} \extran{Saly is growing to be very tall.}
\end{itemize}
\item \anpl{sllollongg} \apl{[]} \defi{to sit close together}
\begin{itemize}
\item \exsen{Ubi komllaebmae ansllollonggallo duliballe.} \extran{The two of them sat close together on that side.}
\end{itemize}
\item \anpl{snameny} \apl{[]} \defi{a funeral tradition, where you take the body to an isolated place and one person will stay with the body and a spirit will come to tell them how they died.}
\item \anpl{somungg} \apl{someny} \defi{to realize}
\item \anpl{suwe tɨngg} \apl{tɨmeny} \defi{to urinate on}
\item \anpl{sämongg} \apl{sämomeny} \defi{to feel}
\begin{itemize}
\item \exsen{Ngäna mermerang ansomägan.} \extran{I feel fine.}
\item \exsen{Obo pätt me gonsomonggän ada obo itrell de dektta llɨtt dägagän.} \extran{He felt in his body that the doctor had stopped his illness.}
\end{itemize}
\item \anpl{taempäg} \apl{taempmeny} \defi{to show, indicate, reveal}
\begin{itemize}
\item \exsen{Bogo mɨnyi tuk me ulle abal känttatt de yantepägän.} \extran{He will show you all a very large bedroom upstairs.}
\item \exsen{Adi gontepägän obo täbatäbe de.} \extran{God revealed his plan.}
\end{itemize}
\item \anpl{tameny} \apl{[]} \defi{to teach, to dicuss, converse, to tell; converse with, speak to}
\begin{itemize}
\item \exsen{Ngäna bam bantemeny walle gllangglla de.} \extran{I'm going to teach you to swim.}
\item \exsen{Yesu ubira yuwog abal ttoen de sanawang eka walle däntamenynegnän.} \extran{Jesus was teaching them many things using parables.}
\end{itemize}
\item \anpl{tamllängg} \apl{tamällmeny} \defi{to press onto something soft, maybe it will break (like the computer)}
\item \anpl{tongg} \apl{[]} \defi{to point}
\begin{itemize}
\item \exsen{Liseng ttang dontogän Rose bom.} \extran{Liseng pointed her finger at Rose.}
\end{itemize}
\item \anpl{torwamängg} \apl{[]} \defi{causative-applicative form of torwam}
\begin{itemize}
\item \exsen{Ngäna bam dantruwamängg.} \extran{I made you lie down.}
\end{itemize}
\item \anpl{trangg} \apl{[]} \defi{to stop}
\item \anpl{trimpeg} \apl{[]} \defi{slide}
\item \anpl{trongg} \apl{tromeny} \defi{to follow}
\begin{itemize}
\item \exsen{Ngäna däbe ddage dantrog do.} \extran{I followed that stream there.}
\end{itemize}
\item \anpl{trungg} \apl{truminy} \defi{to invite, call over, summon}
\begin{itemize}
\item \exsen{Ngänäm ngämo nag dantrugän Upiara we wayati kongkom e.} \extran{My friend invited me to Upiara to collect watermelons.}
\item \exsen{Ngäna bam trungg allan, wiya!} \extran{I am calling you over, come!}
\end{itemize}
\item \anpl{truwamängg} \apl{truwameny} \defi{to lay (someone) down}
\item \anpl{ttaempäg} \apl{ttaempmeny} \defi{to seperate, divorce}
\begin{itemize}
\item \exsen{Ngämi ttaempägmeny dageyo.} \extran{We two are inseparable.}
\end{itemize}
\item \anpl{ttaempäg} \apl{ttaempmeny} \defi{to shake hands with}
\begin{itemize}
\item \exsen{Nagnag a ngämim ttang deyanttepmenyeyo.} \extran{Friends shook hands with us (excl.).}
\end{itemize}
\item \anpl{ttallängg} \apl{[]} \defi{defend}
\item \anpl{ttalängg} \apl{ttalämeny} \defi{to aim}
\begin{itemize}
\item \exsen{Kwalde mäse pakos tupi di gonttalägän sɨmell zuwoe e.} \extran{Kwalde tried aiming the long arrow to shoot the pig.}
\end{itemize}
\item \anpl{ttoengg} \apl{ttoengeny} \defi{to split up, scatter}
\begin{itemize}
\item \exsen{Ibi naeka peyang ttoengeny e dag.} \extran{We will split up in tears.}
\end{itemize}
\item \anpl{ttongg} \apl{sinen} \defi{to give}
\begin{itemize}
\item \exsen{Ngämlle sawe alle giri de nanttog.} \extran{Give me the knife with your left hand.}
\end{itemize}
\item \anpl{ttängkag} \apl{ttängkameny} \defi{to break with force}
\begin{itemize}
\item \exsen{Pakos ulle walle ngäna ddia bo ddäg kutt deyanttäkeg.} \extran{I broke the deer}
\end{itemize}
\item \anpl{tämbag} \apl{tämbameny} \defi{to add}
\item \anpl{tämbameny} \apl{[]} \defi{to instruct}
\begin{itemize}
\item \exsen{Ause da bom däntäbemenyän, "Abo bongo goeg de naemponomeny."} \extran{The old woman instructed him, "You must burn the garden."}
\end{itemize}
\item \anpl{tängg} \apl{[]} \defi{to urinate on, pee on}
\begin{itemize}
\item \exsen{Ekaklle de suwe näntägan.} \extran{He peed on the ground.}
\end{itemize}
\item \anpl{tängg} \apl{[]} \defi{to urinate on, pee on}
\item \anpl{tänggag} \apl{tänggameny} \defi{to make a dog more sensitive to smells by rubbing lemongrass on their nose, to steal}
\item \anpl{tärangg} \apl{tärameny} \defi{to stop, hold back}
\begin{itemize}
\item \exsen{Grace kollba kälsäre de daspunän walle we, be mälla da tämamae angde ikop dägageyo, ubi Grace bom dantärageyo ada, "Kollba de supun a mudan!"} \extran{Grace threw the small fish back into the water, but when all the women saw her, they held her back, saying, "Don't throw back the fish!"}
\end{itemize}
\item \anpl{wändäg} \apl{wändɨminy} \defi{to crowd}
\begin{itemize}
\item \exsen{Yesu adame obo kollmällang de umllang dägnegän oblle gall kälsäre särämbae e, oba lla da ddone obom beyawändägallo.} \extran{This is why Jesus told his disciples to prepare a small boat for him, so that people wouldn't come crowd him.}
\end{itemize}
\item \anpl{yämbäg} \apl{yämbmeny} \defi{to disown, repudiate}
\begin{itemize}
\item \exsen{Ngäna mɨnyi ddone abal bam baembäg.} \extran{I will never disown you.}
\end{itemize}
\item \anpl{zämängg} \apl{zämmeny} \defi{to be in heaps}
\end{itemize}

\section*{Class IV Verbs}
\begin{itemize}
\item \anpl{blab} \apl{blaeb} \defi{to mature, reach puberty}
\begin{itemize}
\item \exsen{Ge nge da ablaban.} \extran{This coconut has matured.}
\item \exsen{Män a zäme ablaban.} \extran{The girl has already reached puberty.}
\end{itemize}
\item \anpl{bädab} \apl{[]} \defi{to shine brightly (of the moon), to dawn, break, dawn}
\begin{itemize}
\item \exsen{Kokta da näbdaban.} \extran{The moon is shining.}
\item \exsen{Ttongo ag a angde däbdabän ge ttoen de ngäna ddobagabira dällätne, ddone ada gotngoenegänän.} \extran{The next morning when it dawn broke, I told this story to others and they laughed a lot.}
\item \exsen{Kikiem a bädab e ekawang pa dan.} \extran{The kikiem is a bird that sings at dawn.}
\end{itemize}
\item \anpl{bällamb} \apl{[]} \defi{ambush}
\item \anpl{bällgab} \apl{bällgaeb} \defi{to open (e.g. eyes, flowers)}
\begin{itemize}
\item \exsen{Däräng täräpang dagirnän ikop saeyang käsre ikop däbällgabän.} \extran{Dog stayed there a long time with eyes closed, then opened his eyes.}
\end{itemize}
\item \anpl{bänamb} \apl{bänaemb} \defi{to open (something folded, e.g. book, mouth)}
\begin{itemize}
\item \exsen{Buk di bämbnaebmam.} \extran{We will open the book.}
\end{itemize}
\item \anpl{dalab} \apl{dalaeb} \defi{to open, pierce, make a hole}
\begin{itemize}
\item \exsen{Obo käp de pop nedelaban.} \extran{He made a hole in the fruit.}
\end{itemize}
\item \anpl{dangkam} \apl{dangkaem} \defi{to lean, to rely on}
\begin{itemize}
\item \exsen{Lla da ttongo lla da bo patme dangkam allan.} \extran{The man is leaning on the other man.}
\item \exsen{Ngäna ngämo mosen patme dangkaemmäll allan ngämingg i.} \extran{I rely on my brother for help.}
\end{itemize}
\item \anpl{daram} \apl{[]} \defi{shine brightly}
\item \anpl{ddänggab} \apl{[]} \defi{to hold, grab (with one's teeth)}
\begin{itemize}
\item \exsen{Sɨmell a ngoi alle ere ada dänddgabän.} \extran{The pig held it with its teeth.}
\end{itemize}
\item \anpl{duab} \apl{duaeb} \defi{to knock over; blow down}
\begin{itemize}
\item \exsen{Däräng a ine de naduaban.} \extran{The dog knocked the water over.}
\item \exsen{Llo de duduaibnegnän a dattkaemnegän.} \extran{Wind blew down the trees and broke them.}
\end{itemize}
\item \anpl{gallab} \apl{gallaeb} \defi{to open (one's mouth)}
\begin{itemize}
\item \exsen{Eka tameny a bongo abo mudan, kälae ttatt gallab a mudan.} \extran{You are not allowed to speak or open your jaw even a little.}
\item \exsen{Abo ttatt de gallab a mudan.} \extran{Don't you say a word.}
\end{itemize}
\item \anpl{gallwab} \apl{gallwaeb} \defi{to prune}
\item \anpl{gollab} \apl{gollaeb} \defi{to pour, flow}
\begin{itemize}
\item \exsen{Ibi kumuddäga ine kutt de dogollaebaebeya.} \extran{We (incl.) poured out three water buckets.}
\item \exsen{Nogollaeb.} \extran{[You] pour it.}
\end{itemize}
\item \anpl{gäbaeb} \apl{gäbnan} \defi{to jump}
\begin{itemize}
\item \exsen{Män a kädkädag walle we mɨnyi bogbaebän.} \extran{The girl will jump into the cold water.}
\end{itemize}
\item \anpl{idab} \apl{idaeb} \defi{to dig out}
\item \anpl{imam} \apl{imaem} \defi{to leave behind; leave without saying goodbye}
\item \anpl{inam} \apl{inaem} \defi{to cover, to weigh down, press down}
\begin{itemize}
\item \exsen{Ekaklle kok de mɨnyi enanae binamän.} \extran{Earth will completely cover the moon (i.e. during a lunar eclipse).}
\item \exsen{Ngäna sɨmell de dinam ekaklle we.} \extran{I weighed the pig down to the ground.}
\end{itemize}
\item \anpl{ingellab} \apl{[]} \defi{to rise up, said of an animal}
\begin{itemize}
\item \exsen{Bogo ingellabang dan.} \extran{He rises up in attention}
\end{itemize}
\item \anpl{kalltam} \apl{kalltaem} \defi{to split}
\begin{itemize}
\item \exsen{Keta da sanador kalltaematt dan.} \extran{A roof post is a split sago stalk.}
\end{itemize}
\item \anpl{kam} \apl{kaem} \defi{to start, begin, to start}
\begin{itemize}
\item \exsen{kam ma ngättma} \extran{place for starting (i.e. starting line)}
\item \exsen{Yogoll a gongkamän.} \extran{The rain started.}
\item \exsen{Obo ma we ibi we kam allan.} \extran{He is starting to go home.}
\item \exsen{Edeb a ine kämbmeny de dängkamän.} \extran{Edeb started diving in the water.}
\end{itemize}
\item \anpl{kam} \apl{[]} \defi{to cut (e.g. meat, skin)}
\begin{itemize}
\item \exsen{Ngäna ngämo tupi di nakaman.} \extran{I cut my index finger.}
\end{itemize}
\item \anpl{kanyam} \apl{kanyaem} \defi{bend}
\item \anpl{kollam} \apl{kollaem} \defi{to stand up}
\begin{itemize}
\item \exsen{Ngäna ma de dakollam.} \extran{I stood the house up.}
\end{itemize}
\item \anpl{kolldab} \apl{[]} \defi{stab}
\item \anpl{kote ttäkam} \apl{[]} \defi{to hang one's head}
\begin{itemize}
\item \exsen{Lla da kote gottkamän.} \extran{The man hung his head.}
\end{itemize}
\item \anpl{käbab} \apl{[]} \defi{to stop, to stop}
\begin{itemize}
\item \exsen{Ngämo moko da käbab e dan.} \extran{I want to stop.}
\item \exsen{Okbab.} \extran{[You] stop.}
\item \exsen{Näkbabeyo!} \extran{[You all] stop it!}
\end{itemize}
\item \anpl{kädab} \apl{kädaeb} \defi{to break, split, to share, split, portion}
\begin{itemize}
\item \exsen{Ngäna nge de dakdab.} \extran{I split the coconut open.}
\item \exsen{Mälla da lla de bun dakdaebaebneyo.} \extran{The women broke the men's skulls.}
\item \exsen{Wayati de nakdab.} \extran{[You] portion the watermelon.}
\item \exsen{Ttängäm dakdaeballo ada ttongdae bo gänyan, ttongdae bo gänyan, do tämamae lla da mullae täräp gogmallo.} \extran{They split the garden like here's one for him, one for him, until everyone got their fair share.}
\end{itemize}
\item \anpl{kätam} \apl{kätaem} \defi{to splash (of an aquatic animal), to flip, do a cartwheel}
\begin{itemize}
\item \exsen{Tärko da mängalae källa gokätaemän.} \extran{The small fish quickly splashed its tail.}
\end{itemize}
\item \anpl{lluwam} \apl{lluwaem} \defi{to incapacitate, make unable}
\begin{itemize}
\item \exsen{Ngänäm wän a nying me nalluwaman Malläm e ibi we.} \extran{This boil on my foot has made me unable to walk to Malam.}
\end{itemize}
\item \anpl{malam} \apl{malaem} \defi{to obey, follow, to accept}
\begin{itemize}
\item \exsen{Lla da obäne eka de dämalaemneyo.} \extran{People used to obey his words.}
\item \exsen{Ngäna bänene kandärmang eka de malam eran.} \extran{I accept your apology.}
\end{itemize}
\item \anpl{modaeb} \apl{[]} \defi{to feel}
\item \anpl{mällam} \apl{mällaem} \defi{to hold; get, grab, catch}
\begin{itemize}
\item \exsen{Rop de namllam ttang me.} \extran{[You] hold the rope in your hand.}
\item \exsen{Ttongo ngäna skul me tresära bin di mällaem eran.} \extran{I hold the position of treasurer at a school.}
\item \exsen{Bogo llo käp de naspunan nag da pate be nag da ddone namllaman.} \extran{She threw the fruit to her friend but they didn't catch it.}
\item \exsen{Kallkäll a damllamän obom.} \extran{He got cold (lit. the cold got him).}
\end{itemize}
\item \anpl{mällkam} \apl{mällkaem} \defi{to bend}
\begin{itemize}
\item \exsen{Kote gomällkamän ngänaeka de gongkamän.} \extran{He hung his head (lit. bent nape) and started to cry.}
\end{itemize}
\item \anpl{mälläng seam} \apl{[]} \defi{to wrinkle up your nose and breathe out, reaction to a bad smell.}
\item \anpl{ngetam} \apl{[]} \defi{to come down, descend}
\begin{itemize}
\item \exsen{Angetam.} \extran{[You] come down.}
\item \exsen{Baet gongetamän matu we.} \extran{The cuscus came down to the ground.}
\end{itemize}
\item \anpl{ngälladab} \apl{[]} \defi{conceive}
\item \anpl{ngänam} \apl{ngänaem} \defi{to understand, recognize, understanding}
\begin{itemize}
\item \exsen{Ubi mɨnyi eka midd de bängnameyo.} \extran{They will understand the meaning.}
\item \exsen{Bibi ddone ke ngänaem amalla?} \extran{Don't you all understand?}
\item \exsen{Bogo gongnamän ada, nagda obom era kuki dägagän.} \extran{He recognized that his friend had deceived him.}
\end{itemize}
\item \anpl{ngänglam} \apl{[]} \defi{to buzz}
\begin{itemize}
\item \exsen{Ngänglamang dan.} \extran{It's buzzing.}
\end{itemize}
\item \anpl{nyärab} \apl{nyäraeb} \defi{to go up, ascend}
\begin{itemize}
\item \exsen{Tutu gonyärabeya.} \extran{We ascended the hill.}
\end{itemize}
\item \anpl{ottam} \apl{[]} \defi{find}
\item \anpl{pallam} \apl{pallaem} \defi{to cut open}
\begin{itemize}
\item \exsen{Obom dapellaemnän.} \extran{He was cutting him open (i.e. for surgery).}
\end{itemize}
\item \anpl{peyam} \apl{peyaem} \defi{to come out from water, surface, to pop up, reappear, resurface}
\begin{itemize}
\item \exsen{Däbe käza da walle amne me gopeyamän.} \extran{That crocodile surfaced in the middle of the pond.}
\item \exsen{Bongo gänyame opeyamalleǃ} \extran{You're here again!}
\end{itemize}
\item \anpl{pisam} \apl{pisaem} \defi{to tear (of something thin), to tear (something thin)}
\begin{itemize}
\item \exsen{Ngämo kaptte da apisaman.} \extran{My garment tore.}
\item \exsen{Ngäna nyäg de napisaman.} \extran{I broke the basket.}
\item \exsen{Ngäna peba de pisam eran.} \extran{I am tearing the paper.}
\end{itemize}
\item \anpl{päddab} \apl{[]} \defi{to grow, to be cooked, be done}
\begin{itemize}
\item \exsen{Tomato da däpddabän.} \extran{The tomato plant grew.}
\item \exsen{Ngäma wätät a zime opddaenegan.} \extran{Our (excl.) food has finished cooking.}
\end{itemize}
\item \anpl{päkam} \apl{[]} \defi{to bend down, keel over (due to sickness)}
\item \anpl{pällkam} \apl{pällkaem} \defi{to split, break}
\begin{itemize}
\item \exsen{Ngäna kakoll de napällkaman.} \extran{I broke the plate.}
\item \exsen{Mälla da täl de dapällkaemaemneyo.} \extran{The women were splitting the bamboo.}
\item \exsen{Ngäna ngämo ikop glas de kamekame napällkaman.} \extran{I accidentally broke my glasses.}
\end{itemize}
\item \anpl{pɨddab} \apl{[]} \defi{to grow}
\item \anpl{saem} \apl{[]} \defi{to babble}
\begin{itemize}
\item \exsen{Bogo saem allan.} \extran{He's babbling.}
\end{itemize}
\item \anpl{torwam} \apl{[]} \defi{to lie down, to lay down}
\begin{itemize}
\item \exsen{tät me torwamang lla} \extran{man lying down on a stretcher}
\item \exsen{Ngäna torwam allan.} \extran{I am lying down.}
\item \exsen{Ngäna obom buntruwam.} \extran{I will lay him down.}
\end{itemize}
\item \anpl{ttaem} \apl{[]} \defi{to pack}
\begin{itemize}
\item \exsen{Lama mängalae za de dättaemnegän.} \extran{Lama quickly packed her things.}
\end{itemize}
\item \anpl{ttalam} \apl{ttalaem} \defi{to split, crack}
\begin{itemize}
\item \exsen{Käza bo ttatt a gottalamän.} \extran{The crocodile}
\end{itemize}
\item \anpl{ttam} \apl{ttaem} \defi{to call, name, to confess, to mention}
\begin{itemize}
\item \exsen{Da bongo mäse dowae me o utale me nänttam, ddone mɨnyi bongllaeyän adawatta obo llan a ttomoll dageyo.} \extran{If you tried to call him, from near or far, he would not answer because he was deaf in his ears.}
\item \exsen{Kaya walle do Upiara ttängämang a ttaem nallo.} \extran{Upiara villagers call me Kaya.}
\item \exsen{Kumddäga pemli de ngäna amim nättaemnegan, ngämi ubim mɨnyi umllang bägaebeya.} \extran{The three families I named, we (excl.) will tell them.}
\item \exsen{Angde bogo gonttamän, llamda da ddone ada mikutt gogon oba pate.} \extran{When he confessed, the old man got very mad at them.}
\end{itemize}
\item \anpl{ttam} \apl{[]} \defi{to appear}
\begin{itemize}
\item \exsen{Yesu ngattong abal e Meri Magdalin pate gottamän.} \extran{Jesus appeared to Mary Magdalene first.}
\end{itemize}
\item \anpl{ttapeyam} \apl{ttapeyaem} \defi{to open (something long, e.g. door, book)}
\item \anpl{ttäkam} \apl{ttäkaem} \defi{to break, snap}
\begin{itemize}
\item \exsen{Obo ttɨle di dattkamän.} \extran{She broke her leg.}
\item \exsen{Ubi ngäma sabi de ttäkam erallo.} \extran{They are breaking our (excl.) law.}
\end{itemize}
\item \anpl{ttällam} \apl{ttällaem} \defi{to pass, hand, to extend, stretch out, reach out, put out}
\begin{itemize}
\item \exsen{Ngäna bäne pate neil de bänttllam.} \extran{I will pass you the nail.}
\item \exsen{Bäne ttang de ttättlemae nänttllam.} \extran{Stretch your arm out straight.}
\end{itemize}
\item \anpl{täbab} \apl{täbaeb} \defi{to watch, look after, patrol}
\begin{itemize}
\item \exsen{sip ikop täbaebag} \extran{shepherd}
\end{itemize}
\item \anpl{täbab} \apl{täbab} \defi{to watch over}
\item \anpl{tägab} \apl{[]} \defi{to turn upside down}
\begin{itemize}
\item \exsen{Ubi obom tutu we dirngäneyo a ingoll dätägabeyo ine gollab e obo bod att.} \extran{They pulled her onto land and turned her upside down so the water would come out of her mouth.}
\end{itemize}
\item \anpl{tämpeyam} \apl{[]} \defi{to abandon, give up}
\begin{itemize}
\item \exsen{Bongo ewatta ke ngänäm nantäpeyamalle?} \extran{Why have you abandoned me?}
\item \exsen{Paelet Yesu bom Rom mäk lla yaba pate däntäpeyamän ddänggaddängge we.} \extran{Pilate gave Jesus up to the Roman soldiers to be crucified.}
\end{itemize}
\item \anpl{täram} \apl{täraem} \defi{to lead, take, carry, collect}
\begin{itemize}
\item \exsen{Kwakmae obom dätramän walle we.} \extran{Kwakmae led him to the water.}
\item \exsen{Ine da obom dätramän do Dowan tutu.} \extran{The water carried him to Dowan mountain.}
\item \exsen{Ubi bibim mɨnyi kwatt e yatraemneyo.} \extran{They will take you all to court.}
\item \exsen{Polis a obom täram erallo daolle.} \extran{The police are taking him away.}
\end{itemize}
\item \anpl{tärpam} \apl{tärpam} \defi{to put across, to crucify}
\begin{itemize}
\item \exsen{Ubi yuwog abal llo de datärpaemaemeyo walle me.} \extran{They put many sticks across the water.}
\item \exsen{Ubi Yesu bom datärpameyo.} \extran{They crucified Jesus.}
\end{itemize}
\item \anpl{tɨram} \apl{tɨraem} \defi{to open}
\begin{itemize}
\item \exsen{Ge ud a täramang daeya, be sisri papekatt dan.} \extran{This door was open, but now it's closed.}
\end{itemize}
\item \anpl{udab} \apl{udaeb} \defi{to disappear, get lost, go missing, to lose}
\begin{itemize}
\item \exsen{Ngäna udab allan.} \extran{I am getting lost.}
\item \exsen{Angde ubi gongoseyo, ubi ikop dägageyo mamam mätta da Meri erem de dibeyän audaban.} \extran{When they came back, they saw that the red yam that Mary had planted was gone.}
\item \exsen{Ubi obom daudabeyo.} \extran{They lost him.}
\end{itemize}
\item \anpl{umaem} \apl{[]} \defi{to gather, to gather, collect}
\begin{itemize}
\item \exsen{Lla da gumaemalle a eka gontemenyalle täre pallall e.} \extran{The people will gather and discuss the feast.}
\item \exsen{Ngäna towall de umaem anggan komlla tum.} \extran{I gathered the grass into two heaps.}
\end{itemize}
\item \anpl{wangam} \apl{wangaem} \defi{to forget}
\begin{itemize}
\item \exsen{Ddone mänyi bam bawengameya.} \extran{We will not forget you.}
\end{itemize}
\end{itemize}

\section*{Class Irregular Verbs}
\begin{itemize}
\item \anpl{ddonddo} \apl{ddonen} \defi{proud, to boast, brag}
\begin{itemize}
\item \exsen{Ngämi bam ge ddobae ddonddo nalla.} \extran{We (excl.) are very proud of you for this.}
\item \exsen{Ubi oba mängall de gonddoneyo.} \extran{They were boasting about their strength.}
\end{itemize}
\item \anpl{ddäddäg} \apl{ddägnan} \defi{edible animal, game, meat, to eat meat; bite, bite (of an animal), hunger for meat}
\begin{itemize}
\item \exsen{Bogo ddäddäg de dokonggän oblle.} \extran{He cut the meat for her.}
\item \exsen{Ddob ddägnan ma gullem a dadeg.} \extran{There are some edible snakes.}
\item \exsen{Ge ddia de däbänya wa däddägaebeya.} \extran{We cut this deer and ate it.}
\item \exsen{Llɨg a ddäddäg de näddägan.} \extran{The boy ate the meat.}
\item \exsen{Käza da ge ttang me daddägän ngänäm.} \extran{The crocodile bit me in this hand.}
\item \exsen{Da bam kungge da naddägän, bongo mɨnyi ttattlle ampug.} \extran{If a spider bites you, you will become very ill.}
\item \exsen{Ddob llayabira komo ddäddäg a malla ttällanenang dan.} \extran{For some people, the centiepede's bite isn't painful.}
\item \exsen{Ngämim ddäddäg abal da deyagnän.} \extran{We (excl.) were very hungry for meat (lit. hunger got us).}
\end{itemize}
\item \anpl{giddoll} \apl{giddollnen} \defi{life, to live, reside, to stay, remain}
\begin{itemize}
\item \exsen{Oba giddoll makäp me ubi ttoen de mermerangae dangesnegneyo, wätät yagnen a,} \extran{In their life, they did all things properly.}
\item \exsen{Bäne giddoll a mɨnyi ada bogon ddobae llayaba za gämällangag.} \extran{Your life will be like this, stealing things from people.}
\item \exsen{Ttongo ause bo bin a Madima, Kinkin ttängäm me giddollag daeya.} \extran{An old woman, her name is Madima, was living in Kinkin village.}
\item \exsen{Män a Daru me dagirnän.} \extran{The girl lived in Daru.}
\item \exsen{Bongo polle ddäg me nagirne, ngäna do amne we balle.} \extran{You stay outside the fence; I will go in the middle.}
\item \exsen{Ende eka da ddone budabän be bagirnän enanae enanae.} \extran{The Ende language will not be lost, but will remain forever.}
\end{itemize}
\item \anpl{ingong} \apl{ingnen} \defi{dance, sing-sing, to dance}
\begin{itemize}
\item \exsen{ingong täräp} \extran{dance time}
\item \exsen{Obo moko da ingong e dan.} \extran{He wants to dance.}
\item \exsen{Ngäna gänyaolle bengne.} \extran{I will dance this way.}
\item \exsen{Ngäna bam ingongang nallan.} \extran{I am making you dance.}
\item \exsen{Lla da ada dänyängnän menae me adawatta bogo ngattong abal gongttägän ttamän ma ngättma we.} \extran{People were dancing like this on the sides because he got to the finish line first.}
\end{itemize}
\item \anpl{itrell} \apl{[]} \defi{disease, illness, sickness, to get hurt}
\begin{itemize}
\item \exsen{Kandärmang säre bablle pätt a ddone abal agan itrell atta.} \extran{Sorry that the illness has made your body so unwell.}
\item \exsen{Da ubi gullem de bänglläbaebneyo, ubi mɨnyi ddone buitrellnegnän.} \extran{If they pick up the snakes, they won't get hurt.}
\end{itemize}
\item \anpl{kättkätt} \apl{[]} \defi{weaving pattern with alternating cross and chevron, to start a new weaving pattern}
\begin{itemize}
\item \exsen{Ibetty tätäm wipellgallagallab alle nyäng de däkättän.} \extran{Yesterday, Ibetty switched from the wipellgallagallab design to a new one.}
\end{itemize}
\item \anpl{nyanyu} \apl{[]} \defi{action, to act, dramatize}
\begin{itemize}
\item \exsen{Bogo gagäll nyanyu di ngasnges eran mällada pate.} \extran{He's doing bad things to his wife.}
\item \exsen{Ge lla ekameny dan, be mɨnyi eka de ttang alle mae nanyunegneyo.} \extran{This man is mute, but just his hands will act out the words.}
\end{itemize}
\item \anpl{papek} \apl{paknen} \defi{dam, blockage; wall, to block; close, to wall, make a wall}
\begin{itemize}
\item \exsen{Ud a papekatt dan.} \extran{The door has been closed.}
\item \exsen{Bogo de papek allan ttang alle.} \extran{He blocked him with his hand.}
\item \exsen{llo ttoe kädnanatt ma paknen e} \extran{a thing that is taken off trees to make walls}
\item \exsen{Oba ma da pall put alle papekatt dan.} \extran{Their house is walled with pall bark.}
\end{itemize}
\item \anpl{tater} \apl{[]} \defi{mat, to spread out}
\begin{itemize}
\item \exsen{Gaem ai tater gollaeb de inen anggan.} \extran{Gaem is weaving many mats.}
\item \exsen{Ubi oba iddpo de däternegeyo dongki bo ddäg toko me.} \extran{They spread out their cloaks over the donkey's back.}
\end{itemize}
\item \anpl{tungg} \apl{tuminy} \defi{support, to support}
\end{itemize}

